---
output:
  pdf_document: default
  html_document: 
      df_print: paged
---

---
header-includes:
 \usepackage{booktabs}
 \usepackage{float}
 \usepackage{graphicx}
 \usepackage{floatrow}
 \usepackage{caption}
 \usepackage{subcaption}
 \usepackage{subfigure}
geometry:
  left=1in,right=1in,top=1in,bottom=1in
---

```{r global options, include=FALSE, message=FALSE}
library(here)
library(here)
library(adnuts)
library(ggplot2)
library(RColorBrewer)
library(doBy)
library(tidyr)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(purrr)
library(stringr)
library(ggplot2)
library(knitr)
library(kableExtra)
options(knitr.kable.NA = '')
knitr::opts_knit$set(root.dir = '../../' )
library(captioner)
library(bibtex)
library(knitcitations)
library(cowplot)
library(magick)
library(bookdown)
library(ggthemes)
library(scales)
library(gridExtra)
library(RColorBrewer)
require(knitr)
library(scales)
require(kfigr) # devtools::install_github("github mkoohafkan/kfigr")
opts_chunk$set(message=FALSE, warning=FALSE)

table_nums <- captioner::captioner(prefix = "Table 4.",levels=1,auto_space = FALSE)
figure_nums <- captioner::captioner(prefix="Figure 4.",levels = 1, auto_space = FALSE)

knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache=FALSE,results="asis")
#clean_cache(clean = TRUE)

thisyr    = 2019


#source("/Users/ingridspies/admbmodels/flatfish/assessments/R/prelims.R")
source("/Users/ingridspies/admbmodels/Katch/assessments/R/read-rep.R")
YFS1=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m1/fm.rep")
YFS2=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m2/fm.rep")
extra_sd2=read.table("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m2/extra_sd.rep",header=TRUE)
extra_sd1=read.table("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m1/extra_sd.rep",header=TRUE)
B0_2=formatC(1000*YFS2$Bzero,format="d",big.mark=",")
B0_1=formatC(1000*YFS1$Bzero,format="d",big.mark=",")

ABC_this=1000*extra_sd2$ABC_HM[1]
ABC_next=1000*extra_sd2$ABC_HM[1]
TotBio_this=extra_sd2$GM_Biom[1]*1000
TotBio_next=extra_sd2$GM_Biom[2]*1000

femcat=YFS2$catage_f/rowSums(YFS2$catage_f)
malcat=YFS2$catage_m/rowSums(YFS2$catage_m)
```

---
title: 'Assessment of the Yellowfin Sole Stock\ in the Bering Sea and Aleutian Islands'

#author: <h4 style="font-style:normal">"Ingrid Spies, Thomas K. Wilderbuer, Daniel G. Nichol, and James Ianelli"</h4> 
output:
  pdf_document: 
  highlight: zenburn
  html_document: default
---


\begin{centering}
\fontfamily{cmr}
\fontsize{10}{12}
  Ingrid Spies, Thomas K. Wilderbuer, Daniel G. Nichol, and James Ianelli\\
\selectfont
  Alaska Fisheries Science Center, National Marine Fisheries Service \\
  National Oceanic and Atmospheric Administration \\ 
  7600 Sand Point Way NE., Seattle, WA 98115-6349 \\
  {`r format(Sys.time(), '%d %B, %Y')`}\\
\end{centering}

#  Executive summary

### Summary of changes in assessment inputs

Relative to last year’s BSAI SAFE report, the
following substantive changes have been made in the BSAI Yellowfin Sole assessment. 

#### Changes in the data

   1.	The `r thisyr` NMFS eastern Bering Sea shelf bottom-trawl survey biomass estimates and standard error were included.
   2. The `r thisyr-1` fishery age composition was included.
   3. The `r thisyr-1` survey age composition was included.
   4. Estimates of the retained and discarded portions of the `r thisyr-1` catch were added.
   5. The estimate of the total catch made through the end of `r thisyr+1` was used. Catch of 150,000 t was assumed for the `r thisyr+1` and `r thisyr+2` projections.

### Changes in the assessment methods

Two models are considered in this assessment. 
   1. The model used in the `r thisyr-1` assessment is referred to as Model 18.1. 
   2. Model 18.1 used the same natural mortality for males and females, M=0.12. A second model is presented (Model 18.2) that used a fixed value for female natural mortality (M=0.12) and allowed male natural mortality to be estimated by within the model. Model 18.2 was the preferred model.

## Summary of Results

A new base model (18.1) and several variants were introduced in 2018, motivated in part by an unexpected 32% decrease in the survey biomass. Model 18.1 includes the survey mean bottom temperature across stations < 100m as a covariate on survey catchability, as in previous years, but adds survey start date as an additional covariate within the model, based on a recent study by Nichol et al. (2018). In `r thisyr` a new model called Model 18.2 that was based on Model 18.1 was introduced that retains female natural mortality fixed at 0.12, but allows the model to estimate male natural mortality.

Although total biomass and spawning biomass have been declining slowly since 1994, spawning biomass estimated by the new model remains high at 1.85 * BMSY. Therefore, Yellowfin Sole continues to qualify for management under Tier 1a. As in recent years, the 1978-2012 age-1 recruitments and the corresponding spawning biomass estimates were used to determine the Tier 1 harvest recommendations. The SSC supports this time period for determining stock productivity and agrees with the authors’ and PT’s recommendations for ABC and OFL under Tier 1a.

This assessment updates last year’s with results and management quantities that are lower than the `r thisyr-1` assessment primarily due to 1) the `r thisyr` survey biomass point estimate is 32% lower than the `r thisyr-1` estimate and 2) the assessment model estimated a slightly lower survey catchability.  Yellowfin Sole continue to be well-above $B_{MSY}$ and the annual harvest remains below the $ABC$ level.  The female spawning stock is in a slow downward trend.  Management quantities are given below for the current base model (Model 14_1) and a new base model (Model 18_1).

\begin{table}[ht]
\centering
\begin{tabular}{lrr|rr}
  \hline
       & \multicolumn{2}{c|}{As estimated or $\mathit{specified}$ } & \multicolumn{2}{c}{As estimated or $\mathit{recommended}$ }  \\
       & \multicolumn{2}{c|}{$\mathit{last}$ year for:}  & \multicolumn{2}{c}{$\mathit{this}$ year for: }               \\
        Quantity & `r thisyr`      &`r thisyr+1`   & `r thisyr+1`      &`r thisyr+2` \\ 
  \hline
	M (natural mortality rate)	&	0.12	&	0.12	&	0.12, 0.135	&	0.12,0.135 \\
Tier	&	1a	&	1a	&	1a	&	1a \\
Projected total (age 6+) biomass (t)	& 2,388,000 t & 2,331,500 t &	`r formatC(extra_sd2$GM_Biom[1]*1000,format="d",big.mark=",")` t	   &	`r formatC(extra_sd2[2,4]*1000,format="d",big.mark=",")` t  \\
Projected female spawning biomass (t) &	827,900 t	&	796,600 t	&		`r formatC(extra_sd2$SSB[1]*1000,format="d",big.mark=",")` t	         &	`r formatC(extra_sd2$SSB[2]*1000,format="d",big.mark=",")` t	 \\
$\:\:\:\:\:\:B_{100\%}$                                 &	1,236,000 t	&	1,236,000 t	&		`r B0_2` t	       &	`r B0_2` t    \\
$\:\:\:\:\:\:B_{MSY\%}$                             &	451,600 t 	&	 451,600 t	&		`r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")` t	     &	`r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")` t		  \\
$F_{OFL}$                             &	0.118	      &	0.118	      &		`r round(extra_sd2$AM_Fmsyr[1],3)`	   &	`r round(extra_sd2$AM_Fmsyr[2],3)`	\\
$maxF_{ABC}$                          & 0.107	      &	0.107	      &	  `r round(extra_sd2$HM_Fmsyr[1],3)`	     &	`r round(extra_sd2$HM_Fmsyr[2],3)`  	\\
$F_{ABC}$                             &	0.107       &	0.107       &		`r round(extra_sd2$HM_Fmsyr[1],3)`           &	`r round(extra_sd2$HM_Fmsyr[2],3)`        \\
$OFL$    	                            &	281,800 t	&	275,100 t	&	  `r formatC(1000*extra_sd2$OFL_AM[1],format="d",big.mark=",")` t          &	`r formatC(1000*extra_sd2$OFL_AM[2],format="d",big.mark=",")` t       \\
$maxABC$    	                        &	255,100 t	&	249,100 t	&		`r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") ` t      &	`r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",") ` t   \\
$ABC$ 	                             &	255,100 t	&	249,100 t	&		`r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") ` t         &	`r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",")` t       \\
\hline
Status	                              &	`r thisyr-2`	      &	 `r thisyr-1`	      &		`r thisyr-1`             &	`r thisyr`          \\
\hline
Overfishing	                          &	No	        &	n/a	      &	No	                       &	n/a                 \\
Overfished	                          &	n/a	        &	No	      &	n/a	                       &	No                  \\
Approaching overfished	              &	n/a	        &	No	      &	n/a	                       &	No                  \\
\hline
\end{tabular}
\end{table}

Projections are based on estimated catches of 150,000 t used in place of maximum $ABC$ for `r thisyr+1` and `r thisyr+2`.

## Responses to SSC and Plan Team Comments on Assessments in General
In this section, we list new or outstanding comments on assessments in general from the last full assessment in `r thisyr-1`. 

## Responses to SSC and Plan Team Comments Specific to this Assessment

1. The SSC encourages further exploration of the way mortality is handled in the model, for example through the use of sex-specific or time-varying mortality and the authors noted that they may be able to explore this more fully in 2019.

_Authors' response_

In the current assessment, a model was explored that used a fixed value for female natural mortality, and allowed male natural mortality to be estimated by within the model (Model 18.2).

2. Given recent changes in the distribution of other species, the SSC encourages authors to explore variability over time in the proportion of the stock that occurs in the Northern Bering Sea. While the model may account for this portion of the stock through the catchability parameter this assumes that the fraction of the biomass occurring in the NBS has not changed substantially. There is little evidence for a change in the NBS biomass within the area that was surveyed in all three years, but this does not account for the possibly large fraction of Yellowfin Sole in nearshore areas that were not surveyed in 2018. The SSC suggests a few avenues to explore possible changes in distribution of Yellowfin Sole, including a comparison of the full NBS survey area between 2010 and 2017, the application of the VAST model to estimate the proportion of Yellowfin Sole in the NBS over time, and an examination of other available data sources, in particular the ADF&G survey in Norton Sound that has been conducted triennially since 1978 and annually since 2017. The SSC encourages the authors to consider approaches for including the substantial biomass of NBS Yellowfin Sole in the model, with the expectations that NBS surveys will be conducted regularly in the future.

_Authors' response_

The distribution of Yellowfin Sole in the eastern Bering Sea is shown in `r figure_nums("NBS",display="cite")`, based on results from the four most recent NBS surveys. There does not appear to be a large shift in distribution since 2010. The distribution of Yellowfin Sole will continue to be monitored and approaches for including the substantial biomass of Yellowfin Sole in the NBS will be considered in future assessments.

# Introduction

The Yellowfin Sole (_Limanda aspera_) is one of the most abundant flatfish species in the eastern Bering Sea (EBS) and currently is the target of the largest flatfish fishery in the world.  They inhabit the EBS shelf and are considered one stock.  Abundance in the Aleutian Islands region is negligible.
Yellowfin Sole are distributed in North American waters from off British Columbia, Canada, (approx. lat. 49$^{\circ}$N) to the Chukchi Sea (about lat. 70$^{\circ}$N) and south along the Asian coast off the South Korean coast in the Sea of Japan (to about lat. 35$^{\circ}$N).  Adults exhibit a benthic lifestyle and occupy separate winter, spawning and summertime feeding distributions on the eastern Bering Sea shelf.  From over-winter grounds near the shelf margins, adults begin a migration onto the inner shelf in April or early May each year for spawning and feeding.  The directed fishery historically occurred from winter through autumn (Wilderbuer et al. 1992).  Yellowfin Sole are managed as a single stock in the BSAI management area as there is presently no direct evidence of stock structure.

# Data      
The data used in this assessment include estimates of total catch, bottom trawl survey biomass estimates and their attendant 95% confidence intervals, catch-at-age from the fishery, and population age composition estimates from the bottom trawl survey.  Weight-at-age and proportion mature-at-age are also available from studies conducted during the bottom trawl surveys.

Data source                                         Year
--------------------------------------------------- ----------------------------   
Fishery catch                                       1954 - `r thisyr`            
Fishery age composition	                            1964 - `r thisyr-1`          
Fishery weight-at-age	                              Avg wt at age from 2008-16 used for 2008-`r thisyr` 
Survey biomass and standard error                   1982 - `r thisyr` 
bottom temperature	                                 1982 - `r thisyr` 
Survey age composition	                             1979 - `r thisyr-1` 
Annual length-at-age and weight-at-age from surveys	1979 - `r thisyr-1` 
Age at maturity	                                    Combined 1992 and 2012 samples
----------------------- 

_Fishery Catch and Catch-at-Age_

This assessment uses fishery catch data from 1954-`r thisyr` (`r table_nums("catch",display="cite")`), and fishery catch-at-age (proportions) from 1964-`r thisyr-1` (`r table_nums("catageprop",display="cite")`, 1975-`r thisyr-1`).  The fishery age composition has always been primarily composed of fish older than 9 years with a large amount of 20+ fish, although the proportion has declined from 90% over age 7 to 70% over age 7 since the 1970's.

_Age Determination_

Yellowfin Sole ages have been determined at the AFSC by using the break and burn method on otoliths collected in surveys and from fisheries since 1979.  In 2016 the age determination methods for Yellowfin Sole were validated using the bomb-produced uptake measurement of 14C method (Kastelle et al. 2016).

_Length and Weight-at-Age_ 

Past assessments of Yellowfin Sole have used sex-specific, time-invariant growth based on the average length-at-age and weight-at-length relationships from the time-series of survey observations summed over all years since 1982.  These weight-at-age estimates were estimated from the von Bertalanffy growth curve.

Parameters of the von Bertalanffy growth curve have been estimated for Yellowfin Sole, by sex, from the trawl survey database as follows: 

Sex     $L_{inf}$  $K$     $t_0$     $n$
------- ---------- ------- --------- ----
Males   33.7       0.161   -0.111    656
Females 37.8       0.137   0.112     709
-----------------------------------------

A sex-specific length-weight relationship was also calculated from the survey database using the power function, 
$Weight(g) =  a*Length(cm)^b$, where $a$ and $b$ are parameters estimated to provide the best fit to the data (Fig. 4.8).  

Sex     $a$       $b$     $n$  
------- --------- ------- -------
Males   0.00854   3.081   2,701
Females 0.0054    3.227   3,662
---------------------------------

These estimates of weight at length were applied to the annual trawl survey estimates of population length at age, by sex, to calculate the weight at each age (Fig. 4.8).  Since the resulting estimates of annual weight-at-age were highly variable for fish older than 11 years, ages 11-20 were smoothed using a five-year average smoothing method for 1982-2018.

Applications of dendrochronology (tree-ring techniques) have been used to develop biochronologies from the otolith growth increments of northern rock sole (_Lepidopsetta polyxystra_), Yellowfin Sole and Alaska plaice (_Pleuronectes quadrituberculatus_) in the eastern Bering Sea. These techniques ensure that all growth increments are assigned the correct calendar year, allowing for estimation of somatic growth by age and year for chronologies that span approximately 25 years (Matta et al. 2010).  The analysis indicated that Yellowfin Sole somatic growth exhibits annual variability and has a strong positive correlation with May bottom water temperature in the Bering Sea (Fig. 4.9).

The relationship between temperature and growth was further explored by reanalyzing Yellowfin Sole growth by age and year.  Length-weight data collected when obtaining otolith (age) samples in RACE surveys (n=7,000 from 1987, 1994 and 1999-2009) also indicate that weight at age exhibits annual variability and is highly correlated with summer bottom water temperature observations with a lag of 2-3 years for the temperature effect to be seen (shown for age 5 fish in figure 4.10) .  These observations were then extended back to 1979 using survey population length-at-age estimates (since weight-at-age is a power function of the length-at-age, Clark et al. 1999, Walters and Wilderbuer 2000). 

In order to incorporate time-varying (year effect on growth) and temperature-dependent growth functions into the age-structured stock assessment model we used the annual observed population mean weight-at-age (time-varying) from the trawl survey. These empirical data indicate good somatic growth correspondence with annual bottom temperature anomalies from 1982-2017 (Fig. 4.11).  Fishery weight at age data available from 2008-2016 were averaged across years for each age to provide updated estimates for the fishery 

_Maturity-at-age_

Maturity information collected from Yellowfin Sole females during the 1992 and 1993 eastern Bering Sea trawl surveys have been used in this assessment for the past 20 years (Table 4.10).  Nichol (1995) estimated the age of 50% maturity at 10.5 years based on the histological examination of 639 ovaries.  Maturity has recently been re-evaluated from a histological analysis of ovaries collected in 2012 (Table 4.10). Results were very similar to the earlier study with only a 2% difference in estimates of Yellowfin Sole female spawning biomass (TenBrink and Wilderbuer 2015).  In addition, the SSC requested that the assessment use a maturity schedule that uses estimates derived from both the 1992 and the 2012 collections (Table 4.10). For Yellowfin Sole sexual maturity occurs well after the age of entry into the fishery.  Yellowfin Sole females are 82% selected to the fishery by age 10 whereas they have been found to be only 40% mature at this age. 

## Fishery     

Yellowfin Sole have annually been caught with bottom trawls on the Bering Sea shelf since the fishery began in 1954 and were overexploited by foreign fisheries in 1959-62 when catches averaged 404,000 t annually (Fig. 4.1, top panel).  As a result of reduced stock abundance, catches declined to an annual average of 117,800 t from 1963-71 and further declined to an annual average of 50,700 t from 1972-77.  The lower yield in this latter period was partially due to the discontinuation of the U.S.S.R. fishery.  In the early 1980s, after the stock condition had improved, catches again increased reaching a peak of over 227,000 t in 1985. 

During the 1980s, there was also a major transition in the characteristics of the fishery.  Yellowfin Sole were traditionally taken exclusively by foreign fisheries and these fisheries continued to dominate through 1984.  However, U.S. fisheries developed rapidly during the 1980s in the form of joint ventures, and during the last half of the decade began to dominate and then take all of the catch as the foreign fisheries were phased out of the EBS.  Since 1990, only domestic harvesting and processing has occurred.  

The management of the Yellowfin Sole fishery changed significantly in 2008 with the implementation of Amendment 80 to the BSAI Fisheries Management Plan.  The Amendment directly allocated fishery resources among BSAI trawl harvesters in consideration of their historic harvest patterns and future harvest needs in order to improve retention and utilization of fishery resources by the non-AFA trawl catcher/processor fleet.  This was accomplished by extending the groundfish retention standards to all H&G vessels and also by providing the ability to form cooperatives within the newly formed Amendment 80 sector.  In addition, Amendment 80 also mandated additional monitoring requirements which included observer coverage on all hauls, motion-compensating scales for weighing samples, flow scales to obtain accurate catch weight estimates for the entire catch, no mixing of hauls and no on-deck sorting.  The partitioning of TAC and PSC (prohibited species catch) among cooperatives has significantly changed the way the annual catch has accumulated (Fig 4.1, bottom panel) and the rate of target catch per bycatch ton. There is now a more even and slow attainment of the annual catch relative to the pre-Amendment 80 fishing behavior.  

Yellowfin Sole are usually headed and gutted, frozen at sea, and then shipped to Asian countries for further processing (AFSC 2016).  The first wholesale value of Alaska Yellowfin Sole totaled $97.8 million in 2014.  From 2016 to 2017 the first wholesale price of all Bering Sea flatfish fisheries increased 16% to $192.9 million t and Yellowfin Sole price increased 19% year over year. In 2010, following a comprehensive assessment process, the Yellowfin Sole fishery was certified under the Marine Stewardship Council environmental standard for sustainable and well-managed fisheries.  The certification also applies to all the major flatfish fisheries in the BSAI and GOA.  The total annual catch (t) since implementation of the MFCMA in 1977 is shown in `r table_nums("catch",display="cite")`.

In 2011, federally permitted vessels using non-pelagic trawl gear whose harvest resulted in flatfish retained catch that was greater than any other retained fishery category were required to use modified trawl gear.  The modifications required the use of elevating devices to raise the section of the trawl warps between the doors and the trawl wing tips by 2.5 inches off the seafloor.  The purpose of the management action was to reduce damage of non-target animals, particularly those that form habitat structure or support other fisheries while not substantially reducing flatfish catch rates or causing gear handling problems (Rose et al. 2010).

In 2011, federally permitted vessels using non-pelagic trawl gear whose harvest results in flatfish retained catch that is greater than any other retained fishery category were required to use modified trawl gear.  The modifications required the use of elevating devices to raise the section of the trawl warps between the doors and the trawl wing tips by 2.5 inches off the seafloor.  The purpose of the management action was to reduce damage of non-target animals, particularly those that form habitat structure or support other fisheries while not substantially reducing flatfish catch rates or causing gear handling problems (Rose et al. 2010).

The 1997 catch of 181,389 t (retained and discards) was the largest since the fishery became completely domestic, but decreased from 1998–2010, averaging 94,004 t (`r table_nums("retdis",display="cite")`).  From 2011-2014 the catch increased, averaging 155,000 t.  The 2013 catch totaled 165,000 t (73% of the ABC), the highest annual catch in the past 19 years.  For `r thisyr`, the catch distribution has been spread out from January through May with the majority coming from four BSAI management areas (509, 513, 514, 517).  As of mid-September `r thisyr`, the fishing season is ongoing.  In order to estimate the total `r thisyr` catch for the stock assessment model, the average proportion of the `r thisyr-8`-`r thisyr-1` cumulative catch attained by the 35th week of the year (mid-September) was applied to the `r thisyr` catch amount at the same time period and results in a `r thisyr` catch estimate of 154,200 t (`r round(100*154200/ABC_this,2)`% of the ABC).  The size composition of the `r thisyr` catch for both males and females, from observer sampling, are shown in Figure 4.2, the catch proportions by month and area are shown in Figure 4.3. Maps of the locations where Yellowfin Sole were caught in `r thisyr`, by month (through mid-September), are shown in Figure 4.4.  The average age of Yellowfin Sole in the `r thisyr-1` catch is estimated at `r round(sum(femcat[(nrow(femcat)-1),]*seq(1,20,1)),2)` and `r round(sum(malcat[(nrow(malcat)-1),]*seq(1,20,1)),2)` years for females and males, respectively.

The time-series of catch in `r table_nums("catch",display="cite")` also includes Yellowfin Sole that were discarded in domestic fisheries from 1987 to the present.  Annual discard estimates were calculated from at-sea sampling (`r table_nums("retdis",display="cite")`).  The rate of discard has ranged from a low of 2% of the total catch in 2012 (and 2015) to 30% in 1992 and is estimated at 3% in the `r thisyr-1` fishery.  The trend has been toward fuller retention of the catch in recent years, and with the advent of the Amendment 80 harvest practices, discarding is at its lowest level since these estimates have become available.  Historically, discarding primarily occurred in the Yellowfin Sole directed fishery, with lesser amounts in the Pacific cod, Pollock, rock sole, flathead sole, and “other flatfish” fisheries (`r table_nums("retdiscat",display="cite")`).

## Survey     

_Survey Biomass Estimates and Population Age Composition Estimates_

Indices of relative abundance available from AFSC surveys showed a major increase in the abundance of Yellowfin Sole during the late 1970s, increasing from 21 kg/ha in 1975 to 51 kg/ha in 1981 (Fig. 4.2 in Bakkala and Wilderbuer 1990).  These increases have also been documented through Japanese commercial pair trawl data and catch-at-age modeling in past assessments (Bakkala and Wilderbuer 1990).

Since 1981, the survey CPUEs have fluctuated widely (Fig. 4.5).  Biomass estimates for Yellowfin Sole from the annual bottom trawl survey on the eastern Bering Sea shelf are shown in `r table_nums("biomCI",display="cite")`, `r table_nums("NBS_bio",display="cite")` and Figure 4.6.  The data show a doubling of survey biomass between 1975 and 1979 with a further increase to over 3.3 million t in 1981.  Total survey abundance estimates fluctuated erratically from 1983 to 1990 with biomass ranging from as high as 3.5 million t in 1983 to as low as 1.9 million t in 1986. Biomass estimates since 1990 indicate an even trend at high levels of abundance for Yellowfin Sole, with the exception of the results from the 1999 and 2000 summer surveys, which were at lower levels.  Surveys from 2001-2005 estimated an increase each year but the estimates since 2006 indicate a stable level with some annual variability.  However, the 2012 estimate is a 19% decrease from 2011 and the 2013 and 2014 surveys have estimated a 17% increase over 2012.  Similarly, there was a 24% decrease from 2014 to 2015 followed by a 48% increase from 2015 to 2016, the highest biomass estimate since 1984. Fluctuations of the magnitude shown between 1980 and 1990, 1998 and 1999, 2008 and 2009, 2011 and 2012, 2014 and 2015, and 2015 and 2016 are unreasonable considering the elements of slow growth and long life span of Yellowfin Sole combined with low to moderate exploitation rate, characteristics which should produce more gradual changes in abundance.

Variability of Yellowfin Sole survey biomass estimates (Fig. 4.6) is in part due to the availability of Yellowfin Sole to the survey area (Nichol 1998).  Yellowfin Sole are known to undergo annual migrations from wintering areas off the shelf-slope break to near shore waters where they spawn throughout the spring and summer months (Nichol, 1995; Wakabayashi, 1989; Wilderbuer et al., 1992).  Exploratory survey sampling in coastal waters of the eastern Bering Sea during early summer indicate that Yellowfin Sole concentrations can be greater in these shallower areas not covered by the standard AFSC survey than in the survey proper.  Commercial bottom trawlers have commonly found high concentrations of Yellowfin Sole in areas such as near Togiak Bay (Low and Narita, 1990) and in more recent years from Kuskokwim Bay to just south of Nunivak Island.  The coastal areas are sufficiently large enough to offer a substantial refuge for Yellowfin Sole from the current survey.  

Over the past 18 years, survey biomass estimates for Yellowfin Sole have shown a positive correlation with shelf bottom temperatures (Nichol, 1998); estimates have generally been lower during cold years.  The 1999 survey, which was conducted in exceptionally cold waters, indicated a decline in biomass that was unrealistic.  The bottom temperatures during the 2000 survey were much warmer than in 1999, and the biomass increased, but still did not approach estimates from earlier years.  Average bottom temperature and biomass both increased again during the period 2001 – 2003, with the 2003 value the highest temperature and biomass observed over the 22 year time series up to that time.  Given that both the 1999 and 2000 surveys were conducted two weeks earlier than previous surveys, it is possible that the time difference may also have also affected the availability of Yellowfin Sole to the survey.  If, for example, the timing of peak Yellowfin Sole spawning in nearshore waters corresponded to the time of the survey, a greater proportion of the population would be unavailable to the standard survey area.  This pattern was observed again in 2009 and 2012 when the temperatures and the bottom trawl survey point estimates were lower.  Summer shelf bottom temperatures in 2012 were the 2nd coldest recorded by the survey and the time-series and resulted in a 19% decline from 2011.  In 2016 the Bering Sea had the highest recorded bottom temperature since measurements began in 1982 and the 2016 estimate of biomass was the highest in 32 years and 48% higher than the 2015 estimate.  The 2017 survey estimate of 2,787,700 t was 3% lower than 2016, but the 2018 estimate of 1,892,925 (another warm year) was down 32% from 2017.
We propose two possible reasons why survey biomass estimates are lower during years when bottom temperatures are low.  First, catchability may be lower because Yellowfin Sole may be less active when cold.  Less active fish may be less susceptible to herding, and escapement under the footrope of survey gear may increase if fish are less active.  Secondly, bottom temperatures may influence the timing of the inshore spawning migrations of Yellowfin Sole and therefore affect their availability to the survey area (Nichol et al. in review).  Because Yellowfin Sole spawning grounds include nearshore areas outside the survey area, availability of fish within the survey area can vary with the timing of this migration and the timing of the survey.  In the case of 2016, a very warm year in the Bering Sea, it appears that a higher portion of the adult biomass was distributed on the shelf (outside of the spawning areas) relative to the average of all previous survey years, indicating earlier spawning migration (Fig 4.7).  

Yellowfin Sole population numbers-at-age estimated from the annual bottom trawl surveys are shown in Table 4.6 and their occurrence in trawl survey hauls and associated collections of lengths and age structures since 1982 are shown in Table 4.7.  Their total tonnage caught in the resource assessment surveys since 1982 are listed in Table 4.8 and also in an appendix table with IPHC survey catches.

_Northern Bering Sea survey_

Trawl survey sampling was extended to the northern Bering Sea in 2010, 2017, 2018, and 2019.  The trawl surveys conducted in 2010, 2017, and 2019 occupied the same areas with similar sampling densities.  The 2018 survey was a reduced effort and only sampled a subset of the northern Bering Sea.  Stations in 2018 were 30 nautical miles apart (instead of 20 nm) and excluded Norton Sound and inshore areas north of Nunivak Island.  For comparison among years (2010, 2017, 2018), biomass estimates were derived by truncating the areal coverage of the 2010 and 2017 surveys to include only the area covered in 2018 that was common to all three surveys, and this was treated as a single stratum (Table 4.5).   This truncated area is 158,286 square kilometers (compared to 200,207 square kilometers in 2010 & 2017).  There was a small increase in the estimate of Yellowfin Sole in the truncated survey area from the 3 surveys. Since Yellowfin Sole fishing is presently prohibited in the northern Bering Sea, the biomass from this area is not included in the stock assessment model. Large shifts in the abundance of Yellowfin Sole into the Bering Sea have not been observed (`r figure_nums("NBS",display="cite")`), but the spatial distribution will continue to be monitored as shifts may occur under future climate change.

# Analytic Approach

## General Model Structure
The abundance, mortality, recruitment and selectivity of Yellowfin Sole were assessed with a stock assessment model using the AD Model Builder language (Fournier et al. 2012; Ianelli and Fournier 1998).  The conceptual model is a separable catch-age analysis that uses survey estimates of biomass and age composition as auxiliary information (Fournier and Archibald 1982).  The assessment model simulates the dynamics of the population and compares the expected values of the population characteristics to the characteristics observed from surveys and fishery sampling programs.  This was accomplished by the simultaneous estimation of the parameters in the model using the maximum likelihood estimation procedure.  The fit of the simulated values to the observable characteristics was optimized by maximizing a log(likelihood) function given some distributional assumptions about the observed data.  
The model starts at age one and fish older than twenty are allowed to accumulate into a plus group. Since the sex-specific weight-at-age for Yellowfin Sole diverges after age of maturity (about age 10 for 40% of the stock) with females growing larger than males, the current assessment model is coded to accommodate the sex-specific aspects of the population dynamics of Yellowfin Sole.  The model allows for the input of sex-specific estimates of fishery and survey age composition and weight-at-age and provides sex-specific estimates of population numbers, fishing mortality, selectivity, fishery and survey age composition and allows for the estimation of sex-specific natural mortality and catchability.  The model retains the utility to fit combined sex data inputs.
The suite of parameters estimated by the model are classified by three likelihood components:

Data component	                           Distributional assumption
----------------------------------------- ---------------------------
Trawl fishery catch-at-age	               Multinomial
Trawl survey population age composition	  Multinomial
Trawl survey biomass estimates and S.E.	  Log-normal
----------------------------------------------------------------

The total log likelihood is the sum of the likelihoods for each data component (Table 4.11).  The likelihood components may be weighted by an emphasis factor, however, equal emphasis was placed on fitting each likelihood component in the Yellowfin Sole assessment except for the catch. The AD Model Builder software fits the data components using automatic differentiation (Griewank and Corliss 1991) software developed as a set of libraries (AUTODIFF C++ library).  Table 4.11 also presents the key equations used to model the Yellowfin Sole population dynamics in the Bering Sea and Table 4.12 provides a description of the variables used in Table 4.11.
Sharp increases in trawl survey abundance estimates for most species of Bering Sea flatfish between 1981 and 1982 indicate that the 83-112 trawl was more efficient for capturing these species than the 400-mesh eastern trawl used in 1975, and 1979-81.  Allowing the model to tune to these early survey estimates would most likely underestimate the true pre-1982 biomass, thus exaggerating the degree to which biomass increased during that period.  Although this underestimate would have little effect on the estimate of current Yellowfin Sole biomass, it would affect the spawner and recruitment estimates for the time-series.  Hence, the pre-1982 survey biomass estimates were omitted from the analysis.

Total mortality $Z$ in the model was modeled as the sum of fishing mortality $F$ and natural mortality $M$, such that total mortality in year $t$ at age $a$ is $Z_{t,a}=F_{t,a}+M.$ 

Fishing mortality at each year and age, $F_{t,a}$, was the product of age-specific fishing gear selectivity $s_a$ and the median year-effect of fishing mortality ${\mu}^F$, with normally distributed error, $$F_{t,a}=s_a{{\mu}^F}e^{{\epsilon}^F_t}, {{\epsilon}^F_t}\backsim N(0,{{\sigma}_F^2}),$$ where ${\epsilon}^F_t$ is the residual year-effect of fishing mortality and ${\sigma}_F$ is the standard deviation of fishing mortality. Age-specific fishing selectivity $s_a$ was calculated using the logistic equation 

$$s_a={\displaystyle\frac{1}{1+e^{(-{\alpha}+age{\beta})}}}.$$

Catch in year $t$ for age $a$ fish $C_{t,a}$ was calculated: $$C_{t,a}={\displaystyle\frac{F_{t,a}}{Z_{t,a}}}(1-e^{Z_{t,a}})N_{t,a},$$ where $N_{t,a}$ is the number of fish at time $t$, age $a$. Total catch in each year $C_t$ was the sum of catch over all ages, $C_t={\sum_a}C_{t,a},$ and the proportion at age in catch was $P_{t,a}={\displaystyle\frac{C_{t,a}}{C_t}}.$

Recruitment from 1956-1975 was modeled as  $N_{t,1}=R_{t}=R_{0}e^{\tau_{t}}, {\tau}_t \backsim\ N(0,\sigma^2_R)$, where $R_0$ is the geometric mean of the modeled age 1 recruitment from 1956-1975, and $\sigma_{R}$ is the standard deviation of recruitment.

Recruitment from 1976-`r thisyr` was determined using the Ricker stock recruitment curve, $$R={\alpha}Se^{-{\beta}S}.$$ Parameters $\alpha$ and $\beta$ were estimated by fitting spawning biomass and recruitment during the period 1976-`r thisyr-6`.

The number of fish in year $t+1$ at age $a$ was the number of fish in the previous year subjected to natural and fishing mortality, $$N_{t+1,a+1}=N_{t,a}e^{-Z_{t,a}}.$$

The number of fish in a "plus group" which included all fish age 20 and older included fish surviving from age 19 as well as those age 20 and higher, $$N_{t+1,A}=N_{t,a}e^{-Z_{t,A-1}}+N_{t,A}e^{-Z_{t,A}}.$$

Spawning biomass was calculated as the product of weight at age and the number of mature females at each age, $$S_t={\sum}N_{t,a}W_{t,a}{\phi}_a,$$ where ${\phi}_a$ is the proportion of mature females at age $a$ and $W_{a,t}$ is the mean body weight in kg of fish age $a$ in year $t$. Survey biomass was assumed to be the product of catchability $q$, survey selectivity $s_a$, and the biomass, $$Biomass_{survey,t}=q{\sum}N_{t,a}W_{t,a}s_a.$$

The model of Yellowfin Sole population dynamics was evaluated with respect to the observations of the time-series of survey and fishery age compositions and the survey biomass trend since 1982. 

## Description of Alternative Models

## Parameters Estimated Outside the Assessment Model
Natural mortality (M) was initially estimated by a least squares analysis where catch at age data were fitted to Japanese pair trawl effort data while varying the catchability coefficient (q) and M simultaneously.  The best fit to the data (the point where the residual variance was minimized) occurred at a M value of 0.12 (Bakkala and Wespestad 1984).  This was also the value which provided the best fit to the observable population characteristics when M was profiled over a range of values in the stock assessment model using data up to 1992 (Wilderbuer 1992).  Since then, natural mortality has been estimated as a free parameter in some of the stock assessment model runs which have been evaluated the past five years.  A natural mortality value of 0.12 is used for both sexes in the base model presented in this assessment.

Yellowfin Sole maturity schedules were estimated from in-situ observations from two studies as discussed in a previous section (Table 4.10).


## Parameter Estimates
(Use the above heading for Tiers 4-6)
List of parameters that are estimated independently of others (e.g., the natural mortality rate, parameters governing the maturity schedule, parameters governing growth [length at age, weight at length or age]—if not estimated inside the assessment model)
Description of how these parameters are estimated (methods do not necessarily have to be statistical; e.g., M could be estimated by referencing a previously published value)

## Parameters Estimated Inside the Assessment Model
The parameters estimated by the model are presented below:

Fishing mortality	 Selectivity	  Survey catchability	    Year-class strength         Spawner-recruit	       Total
------------------ ------------- ----------------------- -------------------------- ----------------------  -----
66	                268	          4	                      105	                       2	                      445
-----------------------------------------------------------------------------------------------

The increase in the number of parameters estimated in this assessment compared to last year (8) can be accounted for by the input of another year of fishery data and the entry of another year class into the observed population, four more sex-specific fishery selectivity parameters and 2 additional catchability parameters.

_Year Class Strengths_

The population simulation specifies the numbers-at-age in the beginning year of the simulation, the number of recruits in each subsequent year, and the survival rate for each cohort as it moves through the population over time using the population dynamics equations given in Table 4.11.

_Selectivity_

Fishery and survey selectivity was modeled separately for males and females using the two parameter formulation of the logistic function (Table 4.11).  The model was run with an asymptotic selectivity curve for the older fish in the fishery and survey, but still was allowed to estimate the shape of the logistic curve for young fish.  The oldest year classes in the surveys and fisheries were truncated at 20 and allowed to accumulate into the age category 20+ years.  A single selectivity curve, for both males and females, was fit for all years of survey data.

Given that there have been annual changes in management, vessel participation and most likely gear selectivity, time-varying fishing selectivity curves were estimated. A logistic equation was used to model fishery selectivity and is a function of time-varying parameters specifying the age and slope at 50% selection,  $\varphi_t$ and $\eta_t$ , respectively.  The fishing selectivity ($S^f$) for age a and year t is modeled as, 

\begin{align}
S^f_{a,t}=[1+e^{\eta_t(a-\varphi_t)}]^{-1},
\end{align}

where $\varphi_t$ and $\eta_t$ are time-varying and partitioned (for estimation) into parameters representing the mean and a vector of deviations (log-scale) conditioned to sum to zero.  The deviations are constrained by a lognormal prior with a variance that was iteratively estimated.  The process of iterating was to first set the variance to a high value (diffuse prior) of $0.5^2$ and estimate the deviations.  The next step was to compare the variability of model estimates.  The variance of the model estimates were then rounded up slightly and fixed for subsequent runs.  The 2016 values were fixed as the average of the 3 most recent years.

_Fishing Mortality_
The fishing mortality rates (F) for each age and year are calculated to approximate the catch weight by solving for F while still allowing for observation error in catch measurement.  A large emphasis (300) was placed on the catch likelihood component to force the model to closely match the observed catch.

_Survey Catchability_
A past assessment (Wilderbuer and Nichol 2001) first examined the relationship between estimates of survey biomass and bottom water temperature.  To better understand how water temperature may affect the catchability of Yellowfin Sole to the survey trawl, catchability was estimated for each year in the stock assessment model as:

\begin{align}
q=e^{-{\alpha}+{\beta}T},
\end{align}

where $q$ is catchability, $T$ is the average annual bottom water temperature anomaly at survey stations less than 100 m, and $\alpha$ and $\beta$ are parameters estimated by the model.  The catchability equation has two parts.  The $e^-\alpha$ term is a constant or time-independent estimate of $q$. The second term, $e^{\beta*T}$  is a time-varying (annual) $q$ which responds to metabolic aspects of herding or distribution (availability) which can vary annually with bottom water temperature.  The result of incorporating bottom temperature to estimate annual $q$ has resulted in an improved fit to the survey (shown in Figure 4.12 for the base model).  

In this assessment we introduce a revised survey catchability model (Model 18_1) where survey start date (expressed as deviation in days (- and +) from the average survey start date of June 4th) and its interaction with annual bottom water temperature is added to the catchability equation as:

\begin{align}
q=e^{-{\alpha}+{\beta}T+{\gamma}S+{\mu}T:S},
\end{align}

where $T$=survey bottom temperature (averaged per year for all stations <100 m), $S$=survey start date, and $T:S$=interaction of $T$ and $S$.  Earlier survey start dates usually encounter colder water and since the timing of the survey start date is positively correlated with bottom water temperature, improvement in fitting the survey biomass estimates can be gained by estimating two new parameters ($\mu$ and $\gamma$).  Akaike information criterion (AIC) were used to determine if the additional variables ($S$ and $T:S$) improved the regression fit.  The improvement in fit was more than offset by the additional two parameters (Nichol et al. In review).

_Spawner-Recruit Estimation_

Annual recruitment estimates from 1978-2012 were constrained to fit a Ricker (1958) form of the stock recruitment relationship as follows:

\begin{align}
R={\alpha}Se^{-{\beta}S},
\end{align}

where $R$ is age 1 recruitment, $S$ is female spawning biomass in metric tons the previous year, and $\alpha$ and $\beta$ are parameters estimated by the model.  The spawner-recruit fitting is estimated in a later phase after initial estimates of survival, numbers-at-age and selectivity are obtained.

# Results

## Model Evaluation

The model evaluation for this stock assessment involved a two-step process.  The first step was to evaluate the productivity of the Yellowfin Sole stock by an examination of which sets of years to include for spawner-recruit fitting (increased from 1978-2010 to 1978-2012 in this assessment).  The second step evaluated various hypothesized states of nature by fitting natural mortality and catchability estimates in various combinations.

The SSC determined in December 2006 that Yellowfin Sole would be managed under the Tier 1 harvest guidelines, and therefore future harvest recommendations would be based on maximum sustainable yield $MSY$ and the associated fishing effort $F_{MSY}$ values calculated from a spawner-recruit relationship.  $MSY$ is an equilibrium concept and its value is dependent on both the spawner-recruit estimates which are assumed to represent the equilibrium stock size-recruitment relationship and the model used to fit the estimates.  In the Yellowfin Sole stock assessment model, a Ricker form of the stock-recruit relationship was fit to various combinations of these data and estimates of $F_{MSY}$ and $B_{MSY}$ were calculated, assuming that the fit to the stock-recruitment data represents the long-term productivity of the stock.  

For this assessment, two different stock-recruitment time-series were investigated:  the full time-series 1955-2012 (Model 14_2) and the post-regime shift era, 1978-2012 (Model 14_1) (Fig. 4.13) (see Joint Plan Team recommendations for September 2012).   Very different estimates of the long-term sustainability of the stock ($F_{MSY}$ and $B_{MSY}$) are obtained depending on which years of stock-recruitment data are included in the fitting procedure (Table 4.13).  When the entire time-series from 1955-2012 was fit, the large recruitments that occurred at low spawning stock sizes in the 1960s and early 1970s determined that the Yellowfin Sole stock was most productive at a smaller stock size with the result that $F_{MSY}$ (0.208) is higher than $F_{35\%}$ ($F_{35\%}$  = 0.17) and $B_{MSY}$ is 314,800 t (Model 14_2).  If we limit the analysis to consider only recruitments which occurred after the well-documented regime shift in 1977 (Model 14_1), a lower value of $F_{MSY}$ is obtained (0.118) and $B_{MSY}$ is 451,600 t.  Table 4.13 indicates that the ABC values from the Model 14_2 harvest scenario for 2019 would be 239,560 t higher than Model 14_1.  Posterior distributions of FMSY for these models indicate that this parameter is estimated with less uncertainty for Model 14_1 resulting in the reduced buffer between ABC and OFL relative to Model 14_2 (9% for Model 14_1 versus 1% for Model 14_2, Table 4.13 and Fig 4.14).

It is important for the Tier 1 calculations to identify which subset of the stock recruitment data is used.  Using the full time series to fit the spawner recruit curve estimates that the stock is most productive at a small stock size.  Thus $MSY$ and $F_{MSY}$ are relatively high values and $B_{MSY}$ is a lower value.  If the stock was productive in the past at a small stock size because of non density-dependent factors (environment), then reducing the stock size to low levels could be detrimental to the long-term sustainability of the stock if the environment, and thus productivity, have changed from the earlier period.  Since observations of Yellowfin Sole recruitment at low stock sizes are not available from multiple time periods, it is uncertain if future recruitment events at low stock conditions would be as productive as during the late 1960s-early 1970s.

Given the uncertainty of the productivity of Yellowfin Sole at low spawning stock sizes, and because the AFSC policy for reference point time-series selection is to use the post 1977 regime shift values unless there is a compelling reason to do otherwise, the productivity of Yellowfin Sole in this assessment is estimated by fitting the 1977-2012 spawner-recruit data in the model (Model 14_1).

The second step in the model evaluation for this assessment entailed the use of a single structural model to consider the uncertainty in the key parameters natural mortality $M$ and catchability.  Model 14_1 has been the preferred model for the past 5 assessments, and operates by fixing $M$ at 0.12 for both sexes and estimating $q$ using the relationship between survey catchability and the annual average water temperature at the sea floor (from survey stations at less than 100 m).  The other models used in the evaluation represented various combinations of estimating $M$ or $q$ as free parameters with different amounts of uncertainty in the parameter estimates (Wilderbuer et al. 2010).  The results are detailed in those assessments and are not repeated here except for the following observations.
The introduction of survey start date as a variable within the $q$ parameter calculation of the stock assessment model (Model 18_1), similarly improved overall model fits to the survey biomass data with the full model ($q=e^{-\alpha+\beta*T+\gamma*S+\mu*T:S}$) providing the best fit to the survey biomass data compared to models where either annual bottom temperature or start date variable (i.e., constant $q$ across years) were not included, or models with only a bottom temperature variable ($q=e^{-\alpha+\beta*T}$) (Nichol et. al 2018). In particular, inclusion of the start date in the model improved model fits for years 1999 to 2003, years for which the more reduced models clearly overestimated survey biomass. (Panel A constant $q$, Panel B bottom temperature only, and Panel C survey start date, bottom temperature and interaction term.)

FIGURE

Given these results and the AIC evidence (Nichol et al. 2018), Model 18_1 is the model of choice for estimating the Yellowfin Sole stock size and management quantities for the 2019 fishing season.  

Experiments examining the bridle efficiency of the Bering Sea survey trawl indicate that Yellowfin Sole are herded into the trawl path from an area between the wing tips of the net and the point where the bridles contact the seafloor (Somerton and Munro 2001).  The herding experiments suggest that the survey trawl vulnerability (a component of catchability) is greater than 1.0.  The likelihood profile of $q$ from the model indicated a small variance with a narrow range of likely values with a low probability of $q$ being equal to the value of 1.0 in a past assessment (Wilderbuer and Nichol 2003).

A model that allows $M$ to be estimated as a free parameter for males with females fixed at 0.12 provided a better fit to the sex ratio estimated from the annual trawl survey age compositions than did the base model (both sexes fixed at $M$ = 0.12).  However, since the population sex ratio annually observed at the time of the survey is a function of the timing of the annual spawning in adjacent inshore areas, it is questionable that providing the best fit to these observations is really fitting the population sex ratio better.  Thus, the model configuration which utilizes the relationship between annual seafloor temperature and survey start date to estimate survey catchability with $M$ fixed at 0.12 for both sexes (Model 18_1) is the preferred model used to base the assessment of the condition of the Bering Sea Yellowfin Sole resource for the 2018 fishing season. 

## Time Series Results 

A brief consideration of the inputs and changes to the assessment methodology relative to last year (Model 14_1) is given before presenting the preferred model results.  Primary updates in going from Model 14_1 to Model 18_1 were the `r thisyr` catch, the fishery and survey age compositions from `r thisyr-1` and the `r thisyr` survey biomass (32% lower than 2017) and standard error estimates. The fishery and survey weights-at-age were also changed in a small amount to include the latest year of data.  In their totality, these changes produced Model 18_1 $ABC$ and $OFL$ estimates for `r thisyr+1` that were 2% lower than the 2017 assessment (Model 14_1) projections for 2019.

As expected, this small increase produced very similar spawner-recruit curves. 

FIGURE
The 2018 overall estimate (1982 – 2018) of trawl survey catchability decreased from 0.9 to 0.88.  This resulted in slightly higher model estimates of population numbers at age and biomass for the time-series back to 1992 relative to last year’s assessment and increased the estimated level of female spawning biomass.  The model results indicate the stock has been in a slowly declining condition since 1994.  The estimates of total biomass and ABC are a bit lower than those used to manage the stock in 2018.  Seven of the past 11 years have had negative bottom temperature anomalies in the Bering Sea but the last four years have been above the mean.  The temperature-dependent $q$ adjustment for 2017 was 0.92.

_Fishing Mortality and Selectivity_

The assessment model estimates of the annual fishing mortality in terms of age-specific annual F and on fully selected ages are given in Tables 4.14 and 4.15, respectively.  The full-selection $F$ has averaged 0.07 over the period of 1978-2018 with a maximum of 0.11 in 1978 and a minimum of 0.04 in 2001.  Model estimated selectivities (Table 4.16, Fig. 4.15) indicate that both sexes of Yellowfin Sole are 50% selected by the fishery at about age 9 and nearly fully selected by age 13, with annual variability.

_Abundance Trends_

The model estimates $q$ at an average value of 0.88 for the period 1982-2018 which results in the model estimate of the 2018 age 2+ total biomass at 2,786,300 t (Table 4.17).  Model results indicate that Yellowfin Sole total biomass (age 2+) was at low levels during most of the 1960s and early 1970s (700,000-1,000,000 t) after a period of high exploitation (Table 4.17, Fig. 4.16, center left panel).  Sustained above average recruitment from 1967-76 combined with light exploitation resulted in a biomass increase to a peak of 3.6 million t by 1985.  The population biomass has since been in a slow decline as the strong 1981 and 1983 year-classes have passed through the population, with only the 1991, 1995 and 2003 year-classes at levels observed during the 1970s.  The present biomass is estimated at 78% of the peak 1985 level.

The female spawning biomass has also declined since the peak in 1994, with a `r thisyr' estimate of 854,800 t (27% decline).  The spawning biomass has been in a gradual decline for the past 22 years and is 36% above the B40% level and 1.9 times the $B_{MSY}$ level (Fig. 4.16).  The model estimate of Yellowfin Sole population numbers at age for all years is shown in Table 4.18 and the resulting fit to the observed fishery and survey age compositions input into the model are shown in the Figure 4.17.  The fit to the trawl survey biomass estimates are shown in Figure 4.16.  Allowing q to be correlated with annual bottom temperature and survey start date provides a better fit to the bottom trawl survey estimates than using a q fixed at the average value (Fig. 4.18).  Table 4.19 lists the numbers of female spawners estimated by the model for all ages and years. The estimated average age of Yellowfin Sole in the population is 6.5 years for males and females.

Both the trawl survey and the stock assessment model indicate that the Yellowfin Sole resource (total biomass) increased during the 1970s and early 1980s to a peak level during the mid-1980s.  The Yellowfin Sole population biomass slowly decreased over the 23 years since the mid-1990s as the majority of year-classes during those years were below average strength.  Average to above average recruitment from 2006 to 2009 is expected to maintain the abundance of Yellowfin Sole at a level above $B_{MSY}$ in the near future.  The stock assessment projection model indicates a mildly decreasing trend in female spawning biomass through 2023 if the fishing mortality rate continues at the same level as the average of the past 5 years (Fig. 4.22).

_Recruitment Trends_

The primary reason for the sustained increase in abundance of Yellowfin Sole during the 1970s and early 1980s was the recruitment of a series of stronger than average year classes spawned in 1967-76 (Figure 4.19 and Table 4.20).  The 1981 year class was the strongest observed (and estimated) during the 47 year period analyzed and the 1983 year class was also very strong.  Survey age composition estimates and the assessment model also estimate that the 1987 and 1988 year classes were average and the 1991 and 1995 year classes were above average.  With the exception of these 4 year classes, recruitment from 15 of the following 19 years estimated from 1984-2005 (since the strong 1983 year-class) were below the 48 year average, which caused the population to gradually decline.  The 2003 year-class has now been observed multiple times in the age compositions and is clearly a strong year class, similar to some of the strong recruitment mentioned above and have contributed to the reservoir of spawning fish in the current population.  In addition, recruitment from 2006-2009 appear also to be average to above average.

_Historical Exploitation Rates_

Based on results from the stock assessment model, annual average exploitation rates of Yellowfin Sole since 1977 ranged from 3% to 7% of the total biomass, and have averaged 4% (Table 4.15).  Posterior distributions of selected parameters from the preferred stock assessment model used in the assessment are shown in Figure 4.20. The values and standard deviations of some selected model parameters are listed in Table 4.21.

_Retrospective Analysis_

A within-model retrospective analysis is also included for the recommended assessment model (Model 18_1) where retrospective female spawning biomass is calculated by working backwards in time dropping data one year at a time and then comparing the “peeled” estimate to the reference stock assessment model used in the assessment (Fig. 4.21).  The resulting pattern from the current assessment model was less than desirable.

Peculiar to the Yellowfin Sole assessment, in comparison to the northern rock sole and Alaska plaice assessments (that have nice patterns), is the large amount of variability in the annual survey biomass assessments for this stock due to the temperature-influenced availability to the survey.  This large variability in the annual estimates can contribute to undesirable patterns since the earlier years are not fitting the same highly variable information as the current year.  

In the 2017 assessment it was demonstrated that low values of Mohn’s rho and desirable retrospective patterns of female spawning biomass were obtainable if lower values of $M$ and $q$ were used relative to the base model.  The Plan Team (and SSC concurred) requested a plot of the model-estimated female spawning biomass trajectory that reduces the retrospective pattern using $M$ fixed at 0.09 and $q$=1.0 on top of the estimated female spawning biomass trajectory with confidence interval from the assessment.  This comparison is plotted below and indicates the retrospective model runs are outside the confidence intervals of the assessment model spawning biomass trajectory for about a 17 year period from 1978-1995.  Otherwise it is within the uncertainty of the assessment model estimate of female spawning biomass. 

FIGURE

The Plan Team recommended that the assessment continue to explore the retrospective patterns in relation to M and q by profiling over a range of combinations of M and q and recording the resulting values of Mohn’s rho and also total likelihood.  This was accomplished using data through 2018 and yielded the following results.

FIGURE

Natural mortality (M) and catchability (q) profile.  Top panel is Mohn’s rho values and bottom panel is log(likelihood).

The best retrospective patterns (lower values of $M$ and $q$ in top panel) did not occur at corresponding best model fit values (single digit numbers in the bottom panel) of $M$ and $q$ (higher values).  The retrospective technique may not always be the best tool for model selection, at least for BSAI Yellowfin Sole as there is tension between model fit and good retrospective pattern over the range of parameterization examined.

Figures of Yellowfin Sole spawning biomass for selected retrospective M-q combinations:
RETROSPECTIVE FIGURES

# Risk assessment table

## Assessment related considerations

The BSAI Yellowfin Sole assessment is based on surveys conducted annually on the EBS shelf from 1982-2018 (no skipped years).  Fish ages, derived from otoliths collected during the surveys and the fishery to calculate annual estimates of population and fishery age composition, have been validated.  The assessment model exhibits good fits to all compositional and abundance data and converges to a single minima in the likelihood surface.  Recruitment estimates track strong year classes that are consistent with the data.  
The retrospective pattern from the current assessment model was less than desirable and has been the subject of some concern for the assessment.  Peculiar to the Yellowfin Sole assessment, in comparison to the northern rock sole and Alaska plaice assessments (that have nice patterns), is the large amount of variability in the annual survey biomass assessments for this stock due to the temperature-influenced availability to the survey.  This large variability in the annual estimates can contribute to undesirable patterns since the earlier years are not fitting the same highly variable information as the current year.  
In the 2017 assessment it was demonstrated that low values of Mohn’s rho and desirable retrospective patterns of female spawning biomass were obtainable if lower values of M and q were used relative to the base model.  However, the best retrospective patterns do not occur at the corresponding values of best overall model fit of M and q (which occur at higher levels of M and q).  The retrospective technique may not always be the best tool for model selection, at least for BSAI Yellowfin Sole as there is tension between model fit and good retrospective pattern over the range of parameterization examined.

## Population dynamics considerations

Stock assessment model results indicate that Yellowfin Sole total biomass (age 2+) was at low levels during most of the 1960s and early 1970s (700,000-1,000,000 t) after a period of high exploitation.  Sustained above average recruitment from 1967-76 combined with light exploitation resulted in a biomass increase to a peak of 3.6 million t by 1985.  The population biomass has since been in a slow decline as the strong 1981 and 1983 year-classes have passed through the population, with only the 1991, 1995 and 2003 year-classes at levels observed during the 1970s, although the 2006-2010 year-classes all look about average according to the 2018 stock assessment.  The present biomass is estimated at 78% of the peak 1985 level and is at twice the level of BMSY.  Projections indicate that the FSB will remain well-above the BMSY level through 2028.  Population dynamics are not a concern for this assessment.

## Environmental/ecosystem considerations

It is unknown how particular environmental factors affect the sustainability of Yellowfin Sole.  It has been demonstrated that the migration to spawning grounds and subsequent spawning occurs sooner in warm years.  Somatic growth has also been shown to increase in warmer temperatures. The prey field for this species consists primarily of polycheate worms and clam siphons, species whose distribution and abundance patterns relative to a changing environment are unknown. It is also unknown what effect environmental variability has on the distribution of Yellowfin Sole predators.  Given the high level of sustained abundance of this species, there is a low level of concern for a variable environment, at least for the levels observed over the past 40 years.
These results are summarized in the table below.

## Fishery performance considerations
Recent surveys of the northern Bering sea have not indicated a large shift in the spatial distribution of the eastern Bering Sea stock of Yellowfin Sole. If the stock moves northward out of the eastern Bering Sea under climate change into untrawlable areas in the northern Bering sea, then fisheries would be unable to target the stock in the untrawlable zone. A NOAA Coastal and Oceans Climate Applications proposal will be submitted examine the implications to the fishery and the region of the northern Bering Sea if the stock of Yellowfin Sole shifts northward. At the current time, fishery CPUE is not showing a contrasting pattern from the stock biomass trend, unusual spatial pattern of fishing, or changes in the percent of TAC taken, changes in the duration of fishery openings.


## Harvest Recommendations
```{r,echo=FALSE}
library(knitr)

thisyr    = 2019
```


_Scenario Projections and Two-Year Ahead Overfishing Level_

In addition to the seven standard harvest scenarios, Amendments 48/48 to the BSAI and GOA Groundfish Fishery Management Plans require projections of the likely OFL two years into the future.  The `r thisyr` numbers at age from the stock assessment model are projected to 2019 given the 2018 catch and then a 2019 catch of 150,000 t is applied to the projected 2019 population biomass to obtain the 2020 $OFL$. 

The SSC has determined that Yellowfin Sole qualify as a Tier 1 stock and therefore the `r thisyr+1`  $ABC$ is calculated using Tier 1 methodology. The Tier 1 harvest level is calculated as the product of the harmonic mean of $F_{MSY}$ and the geometric mean of the `r thisyr+1`  biomass estimate.

The geometric mean of the `r thisyr+1`  biomass estimate, $B_{gm}$, is estimated using the equation  $$B_{gm}=e^{ln(B)-(cv^2/2)},$$ 
where  $B$ is the point estimate of the `r thisyr+1`  biomass from the stock assessment model and $cv^2$ is the coefficient of variation of the point estimate (a proxy for sigma). The harmonic mean of $F_{MSY}$, $F_{har}$ is estimated as $F_{har}=e^{ln(F_{MSY}-(ln(sd^2)/2)}$, where $F_{MSY}$ is the peak mode of the $F_{MSY}$ distribution and $sd^2$ is the square of the standard deviation of the $F_{MSY}$ distribution.

In 2006 the SSC selected the 1978-2001 data set for the Tier 1 harvest recommendation.  Using this approach again for the `r thisyr+1` harvest (now the 1978-2012 time-series) recommendation (Model 18_1 in Table 4.13), the $F_{ABC}$ =  $F_{harmonic_mean}$ = 0.107. The estimate of age 6+ total biomass for `r thisyr+1` is 2,462,440 t. The calculations outlined above give a Tier 1 ABC harvest recommendation of 263,200 t and an $OFL$ of 289,900 t for `r thisyr+1`.  This results in a 9% (26,200 t) buffer between $ABC$ and $OFL$. The $ABC$ value is 5% lower than last year, primarily due to the 32% decline in the survey estimate from `r thisyr-1` to `r thisyr`. 

The stock assessment analysis must also consider harvest limits, usually described as overfishing fishing mortality levels with corresponding yield amounts. Amendment 56 to the BSAI FMP sets the Tier 1 harvest limit at the $F_{MSY}$ fishing mortality value.  The overfishing fishing mortality values, ABC fishing mortality values and their corresponding yields are given as follows:

Harvest level	                            F value      `r thisyr+1`  Yield    
----------------------------------------- ------------ --------------------
Tier 1 $F_{OFL}$ = $F_{MSY}$              0.117        289,900 t         
Tier 1 $F_{ABC}$ = $F_{harmonic_mean}$    0.107        263,200 t
----------------------------------------------------------------

_Status Determination_

A standard set of projections is required for each stock managed under Tiers 1, 2, or 3 of Amendment 56.  This set of projections encompasses seven harvest scenarios designed to satisfy the requirements of Amendment 56, the National Environmental Policy Act, and the 1977 Magnuson-Stevens Fishery Conservation and Management Act (MSFCMA).

For each scenario, the projections begin with the vector of `r thisyr` numbers at age estimated in the assessment.  This vector is then projected forward to the beginning of `r thisyr+1` using the schedules of natural mortality and selectivity described in the assessment and the best available estimate of total (year-end) catch for `r thisyr`.  In each subsequent year, the fishing mortality rate is prescribed on the basis of the spawning biomass in that year and the respective harvest scenario.  In each year, recruitment is drawn from an inverse Gaussian distribution whose parameters consist of maximum likelihood estimates determined from recruitments estimated in the assessment.  Spawning biomass is computed in each year based on the time of peak spawning and the maturity and weight schedules described in the assessment.  Total catch is assumed to equal the catch associated with the respective harvest scenario in all years.  This projection scheme is run 1,000 times to obtain distributions of possible future stock sizes, fishing mortality rates, and catches.

Five of the seven standard scenarios will be used in an Environmental Assessment prepared in conjunction with the final SAFE.  These five scenarios, which are designed to provide a range of harvest alternatives that are likely to bracket the final TAC for `r thisyr+1`, are as follows (max $F_{ABC}$ refers to the maximum permissible value of $F_{ABC}$ under Amendment 56):

* Scenario 1:  In all future years, $F$ is set equal to max $F_{ABC}$.  (Rationale:  Historically, $TAC$ has been constrained by ABC, so this scenario provides a likely upper limit on future TACs.)
* Scenario 2:  In all future years, F is set equal to a constant fraction of max $F_{ABC}$, where this fraction is equal to the ratio of the FABC value for 2019 recommended in the assessment to the max $F_{ABC}$ for `r thisyr+1`.  (Rationale:  When $F_{ABC}$ is set at a value below max $F_{ABC}$, it is often set at the value recommended in the stock assessment.)
* Scenario 3:  In all future years, $F$ is set equal to the 2014-`r thisyr` average $F$.  (Rationale:  For some stocks, $TAC$ can be well below $ABC$, and recent average F may provide a better indicator of $F_{TAC}$ than $F_{ABC}$.)
* Scenario 4:  In all future years, the upper bound on $F_{ABC}$ is set at F60%. (Rationale: This scenario
provides a likely lower bound on $F_{ABC}$ that still allows future harvest rates to be adjusted
downward when stocks fall below reference levels.)
* Scenario 5:  In all future years, $F$ is set equal to zero.  (Rationale:  In extreme cases, $TAC$ may be set at a level close to zero.)

Two other scenarios are needed to satisfy the MSFCMA’s requirement to determine whether a stock is currently in an overfished condition or is approaching an overfished condition.  These two scenarios are as follow (for Tier 3 stocks, the $MSY$ level is defined as $B_{35\%}$):

* Scenario 6:  In all future years, F is set equal to $F_{OFL}$.  (Rationale:  This scenario determines whether a stock is overfished.  If the stock is expected to be above its $MSY$ level in 2016 and above its $MSY$ level in 2030 under this scenario, then the stock is not overfished.)
* Scenario 7:  In `r thisyr+1` and `r thisyr+2`, $F$ is set equal to max $F_{ABC}$, and in all subsequent years, F is set equal to FOFL.  (Rationale:  This scenario determines whether a stock is approaching an overfished condition.  If the stock is expected to be above its $MSY$ level in `r thisyr+13` under this scenario, then the stock is not approaching an overfished condition.)

Simulation results shown in Table 4.22 indicate that Yellowfin Sole are not currently overfished and are not approaching an overfished condition.  The projection of Yellowfin Sole female spawning biomass through `r thisyr+13` is shown in Figure 4.22 and a phase plane figure of the estimated time-series of Yellowfin Sole female spawning biomass relative to the harvest control rule is shown in Figure 4.23.

Year	 Catch	    SSB	         Geom. mean 6+ biomass	    ABC	            OFL
----- --------- ------------ ------------------------- --------------- ---------
2019	 150,000	  850,600  	   2,462,400	                263,200  	      290,000
2020	 150,000	  821,500	     2,411,700	                257,800  	      284,000
-------------------------------------------------------------

# Ecosystem Considerations
(Authors are encouraged to use information contained in the Ecosystem Considerations chapter to assist them in developing stock-specific analyses and to recommend new information for inclusion in future versions of the Ecosystem Considerations chapter.  Time series currently contained in the Ecosystem Considerations chapter may simply be referenced rather than duplicated here.  In cases where stock-specific time series or relationships are used, this information should be included here rather than in the Ecosystem Considerations chapter.)

## Ecosystem Effects on the Stock
The following factors should be discussed:
Prey availability/abundance trends (historically, in the present, and in the foreseeable future).  These prey trends could affect growth or survival of a target stock. 

1)	Predator population trends (historically, in the present, and in the foreseeable future).  These trends could affect stock mortality rates over time.
2)	Changes in habitat quality (historically, in the present, and in the foreseeable future).  Changes in the physical environment such as temperature, currents, or ice distribution could affect stock migration and distribution patterns, recruitment success, or direct effects of temperature on growth.

## Fishery Effects on the Ecosystem 
The following factors should be discussed:

1)	Fishery-specific contribution to bycatch of prohibited species, forage (including herring and juvenile pollock), HAPC biota (in particular, species common to the target fishery), marine mammals, birds, and other sensitive non-target species (including top predators such as sharks, expressed as a percentage of the total bycatch of that species.
2)	Fishery-specific concentration of target catch in space and time relative to predator needs in space and time (if known) and relative to spawning components.
3)	Fishery-specific effects on amount of large-size target fish.
4)	Fishery-specific contribution to discards and offal production.
5)	Fishery-specific effects on age at maturity and fecundity of the target species.
6)	Fishery-specific effects on EFH non-living substrate (using gear specific fishing effort as a proxy for amount of possible substrate disturbance). 

# Data Gaps and Research Priorities
List areas where a significant improvement in the amount of available information would likely result in a significant improvement in the quality of the assessment and the estimates of critical parameters.

# Literature Cited

Alaska Fisheries Science Center.  2016. Wholesale market profiles for Alaska groundfish and crab fisheries.  134 p. Alaska Fish. Sci. Cent., NOAA, Natl. Mar. Fish. Serv. 7600 Sand Point Way NE. Seattle, WA.

Bakkala, R. G. and V. Wespestad.  1984.  Yellowfin sole.  In R. G. Bakkala and L. resources of the eastern Bering Sea and Aleutian Islands region in 1983, p. 37-60. U.S. Dep. Commer., NOAA Tech. Memo.  NMFS F/NWC-53.

Bakkala, R. G., V. Wespestad, and L. Low.  1982.  The yellowfin sole (Limanda aspera) resource of the eastern Bering Sea--its current and future potential for commercial fisheries.  U.S. Dep. Commer., NOAA Tech. Memo.  NMFS F/NWC-33, 43p.

Bakkala, R. G., and T. K. Wilderbuer.  1990.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands     Region as Projected for 1990, p. 60-78.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak 99510.

Clark, W. G., Hare, S. R., Parms, A. M., Sullivan, P, J., Trumble, R. J.  1999.  Decadal changes in growth and recruitment of Pacific halibut (Hipplglossus stenolepis).  Can. J. fish. Aquat. Sci. 56, 242-252.

Fournier, D. A., H.G. Skaug, J. Ancheta, J. Ianelli, A. Magnusson, M. N. Maunder, A. Nielsen, and J. Sibert.  2012.  AD Model Builder: using automatic differentiation for statistical inference of highly parameterized complex nonlinear models.  Optim. Methods Softw. 27:233-239.

Fournier, D. A. and C.P. Archibald.  1982.  A general theory for analyzing catch-at-age data.  Can. J. Fish Aquat. Sci.  39:1195-1207.

Greiwank, A. and G. F. Corliss (eds) 1991.  Automatic differentiation of algorithms: theory, implementation and application.  Proceedings of the SIAM Workshop on the Automatic Differentiation of Algorithms, held Jan. 6-8, Breckenridge, CO. Soc. Indust. And Applied Mathematics, Philadelphia.

Haflinger, K.  1981.  A survey of benthic infaunal communities of the Southeastern Bering Sea shelf.  In Hood and Calder (editors) The Eastern Bering Sea Shelf: Oceanography and Resources, Vol. 2. P. 1091-1104.  Office Mar. Pol. Assess., NOAA. Univ. Wash. Press, Seattle, Wa 98105.

Ianelli, J. N. and D. A. Fournier.  1998.  Alternative age-structured analyses of the NRC simulated stock assessment data.  In Restrepo, V. R. [ed.]  Analyses of simulated data sets in support of the NRC study on stock assessment methods.  NOAA Tech. Memo.  NMFS-F/SPO-30.  96 p.

Kastelle, C., T. Helser, S. Wischniowski, T. Loher, B. Geotz and L. Kautzi.  2016. Incorporation of bomb-produced 14C into fish otoliths: A novel approach for evaluating age validation and bias with an application to yellowfin sole and northern rockfish. Ecological modeling 320 (2016) 79-91.

Low, L. and R.E. Narita.  1990.  Condition of groundfish resources in the Bering Sea-Aleutian Islands region as assessed in 1988.  U.S. Dep. Commer., NOAA Tech. Memo.  NMFS-F/NWC-178, 224 p.

Matta, M. E., B. A. Black and T. K. Wilderbuer.  2010.  Climate-driven synchrony in otolith growth-increment chronologies for three Bering Sea flatfish species.  MEPS, Vol. 413:137-145, 2010.

Nichol, D. R .  1995.  Spawning and maturation of female yellowfin sole in the eastern Bering Sea.  In Proceedings of the international flatfish symposium, October 1994, Anchorage, Alaska, p. 35-50.  Univ. Alaska, Alaska Sea Grant Rep. 95-04.

Nichol, D. R.  1998.  Annual and between sex variability of yellowfin sole, Pleuronectes asper, spring-summer distributions in the eastern Bering Sea.  Fish. Bull., U.S. 96: 547-561.

Nichol, D.G., Kotwicki, S., Wilderbuer, T.K., Lauth, R.R. and Ianelli, J.N., 2019. Availability of yellowfin sole Limanda aspera to the eastern Bering Sea trawl survey and its effect on estimates of survey biomass. Fisheries Research, 211, pp.319-330.

Ricker, W. E.  1958.  Handbook of computations for biological statistics of fish populations.  Bull. Fish. Res. Bd. Can.,  (119) 300 p.

Rose, C. S., J. R. Gauvin and C. F. Hammond.  2010. Effective herding of flatfish by cables with minimal seafloor contact.  Fishery Bulletin 108(2):136-144.

Somerton, D. A. and P. Munro.  2001.  Bridle efficiency of a survey trawl for flatfish.  Fish. Bull. 99:641-652 (2001).

TenBrink, T. T. and T. K. Wilderbuer.  2015.  Updated maturity estimates for flatfishes (Pleuronectidae) in the eastern Bering Sea, with notes on histology and implications to fisheries management.  Coastal and Marine Fisheries: Dynamics, Management and Ecosystem Science. O:1-9. 2015. DOI: 10.1080/19425120.2015.1091411.

Wakabayashi, K.   1989.  Studies on the fishery biology of yellowfin sole in the eastern Bering Sea. [In Jpn., Engl. Summ.] Bull. Far Seas Fish. Res. Lab. 26:21-152.

Wakabayashi, K., R. Bakkala, and L. Low.  1977.  Status of the yellowfin sole resource in the eastern Bering Sea through    1976.  Unpubl. manuscr., 45p.  Northwest and Alaska Fish. Cent., Natl. Mar. Fish. Serv., NOAA, 7600 Sand Point Way N.E., Bin C 15700, Seattle, Wa  98115.

Walters, G. E. and T. K. Wilderbuer.  2000.  Decreasing length at age in a rapidly expanding population of northern rock sole in the eastern Bering Sea and its effect on management advice.  Journal of Sea Research 44(2000)17-26.

Wilderbuer, T. K.  1992.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands     Region as Projected for 1993, chapter 3.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak 99510.

Wilderbuer, T. K.  1993.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands Region as Projected for 1994, chapter 3.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak  99510.

Wilderbuer, T. K. and D. Nichol.  2001.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands Region as Projected for 2004, chapter 4.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak  99510.

Wilderbuer, T. K. and D. Nichol.  2003.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands Region as Projected for 2004, chapter 4.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak  99510.

Wilderbuer, T. K. D. G. Nichol, and J. Ianelli.  2010.  Yellowfin sole.  In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands Region as Projected for 2011, chapter 4.  North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, Ak  99510.

Wilderbuer, T.K., G.E. Walters, and R.G. Bakkala   1992.  Yellowfin sole, _Pleuronectes aspera_, of the eastern Bering Sea: biological characteristics, history of exploitation, and management.  Mar Fish. Rev. 54(4):1-18.


```{r, echo=FALSE, hierarchy}
table_nums <- captioner::captioner(prefix = "Table 4.",levels=1,auto_space = FALSE)
figure_nums <- captioner::captioner(prefix="Figure 4.",levels = 1, auto_space = FALSE)












tab.11_cap <- table_nums(name = "tab_11", caption = "Key equations used in the population dynamics model.")

tab.12_cap <- table_nums(name = "tab_12", caption = "Variables used in the population dynamics model.")

tab.13_cap <- table_nums(name = "tab_13", caption = "Models evaluated for stock productivity in the 2018 stock assessment of Yellowfin Sole.")

tab.14_cap <- table_nums(name = "tab_14", caption = "Model estimates of annual average fishing mortality for male and female Yellowfin Sole.")

tab.15_cap <- table_nums(name = "tab_15", caption = "Model estimates of Yellowfin Sole full selection fishing mortality and exploitation rate (catch/total biomass).")

tab.16_cap <- table_nums(name = "tab_16", caption = "Model estimates of Yellowfin Sole age-specific selectivities for the survey and fishery (ages 4 to 20 from left to right). ")

tab.17_cap <- table_nums(name = "tab_17", caption = "Model estimates of Yellowfin Sole age 2+ total biomass (t) and begin-year female spawning biomass (t) from the 2017 and 2018 stock assessments.")

tab.18_cap <- table_nums(name = "tab_18", caption = "Model estimates of Yellowfin Sole population numbers at age (billions) for 1954-2018.")

tab.19_cap <- table_nums(name = "tab_19", caption = "Model estimates of the number of female spawners (millions) 1964-2018.")

tab.20_cap <- table_nums(name = "tab_20", caption = "Model estimates of Yellowfin Sole age 5 recruitment (millions) from the 2017 and 2018 stock assessments.")

tab.21_cap <- table_nums(name = "tab_21", caption = "Selected parameter estimates and their standard deviation from the preferred stock assessment model in 2018.")

tab.22_cap <- table_nums(name = "tab_22", caption = "Projections of Yellowfin Sole female spawning biomass (1,000s t), catch (1,000s t) and full selection fishing mortality rate for seven future harvest scenarios.")

tab.23_cap <- table_nums(name = "tab_23", caption = "Catch and bycatch (t) of other BSAI target species in the Yellowfin Sole directed fishery from 1992-2016 estimated from a combination of regional office reported catch and observer sampling of the catch.")

tab.24_cap <- table_nums(name = "tab_24", caption = "Estimated non-target species catch (t) in the Yellowfin Sole fishery, 2003-2018 (PSC not included).")

tab.25_cap <- table_nums(name = "tab_25", caption = "Yellowfin Sole TAC and ABC levels, 1980- 2018")

tab.26_cap <- table_nums(name = "tab_26", caption = "Catch and bycatch (t) of other BSAI target species in the Yellowfin Sole directed fishery from 1992-2016 estimated from a combination of regional office reported catch and observer sampling of the catch.")

tab.27_cap <- table_nums(name = "tab_27", caption = "Catch and bycatch (t) of other BSAI target species in the Yellowfin Sole directed fishery from 1992-2016 estimated from a combination of regional office reported catch and observer sampling of the catch.")

tab.28_cap <- table_nums(name = "tab_28", caption = "Catch and bycatch (t) of other BSAI target species in the Yellowfin Sole directed fishery from 1992-2016 estimated from a combination of regional office reported catch and observer sampling of the catch.")

fig.1_cap = figure_nums(name="fig_1",caption = paste("Yellowfin Sole annual total catch (1,000s t) in the Eastern Bering Sea from 1954-",thisyr," (upper panel). Yellowfin Sole annual cumulative catch by month and year (non CDQ) 2003-",thisyr-1," (lower panel).",sep="")) 

fig.2_cap = figure_nums(name="fig_2",caption = paste("Size composition of the Yellowfin Sole catch in ",thisyr,"(through mid-September), by subarea and total.
                                                     "))

fig.3_cap = figure_nums(name="fig_3",caption = paste("Yellowfin Sole catch proportion by area (upper panel) and by month (lower panel) in the Eastern Bering Sea in ",thisyr,"."))


fig.5_cap = figure_nums(name="fig_5",caption = "Yellowfn sole CPUE (catch per unit effort in kg/ha) from the annual Bering Sea shelf trawl surveys, 1982-2018.")

fig.6_cap = figure_nums(name="fig_6",caption = paste("Annual eastern Bering Sea bottom trawl survey biomass point-estimates and 95% confidence intervals for Yellowfin Sole, 1982-",thisyr,".
                                                     ",sep=""))

fig.7_cap = figure_nums(name="fig_7",caption = "Difference between the 1985-2016 average trawl survey CPUE for Yellowfin Sole and the 2017 survey CPUE.  Open circles indicate that the magnitude of the catch was greater in 2017 than the long-term average, closed circles indicate the catch was greater in the long-term average than in 2017.")

fig.8_cap = figure_nums(name="fig_8",caption = "Average Yellowfin Sole weight-at-age (g) from trawl survey observations.
                        ")

fig.9_cap = figure_nums(name="fig_9",caption = "Master chronology for Yellowfin Sole and time series of mean summer bottom temperature and May sea surface temperature for the southeastern Bering Sea (Panel A).  All data re normalized to a mean of 0 and standard deviation of 1.  Correlations of chronologies with bottom temperature and sea surface temperature are shown in panels B and C, respectively (Matta et al. 2010).")

fig.10_cap = figure_nums(name="fig_10",caption = "Yellowfin Sole length-at-age anomalies, for males and females, and bottom temperature anomalies.  Correspondence in these residuals is apparent with a 2-3 year lag effect from the mid-1990s to 2017.  Late 1980s and early 1990s pattern may be a density-dependent response in growth from the large 1981 and 1983 year-classes.")

fig.11_cap = figure_nums(name="fig_11",caption = "Table of temperature anomalies (second row at top as bars) and observed values by age and year. Shadings within the matrix reflects relative weight-at-age (within a row) with darker red heavier than average.")

fig.12_cap = figure_nums(name="fig_12",caption = "Average bottom water temperature from stations less than or equal to 100 m in the Bering Sea trawl survey (bars) and the stock assessment model estimate of q for each year 1982-2017.")

fig.13_cap = figure_nums(name="fig_13",caption = "Fit of the Ricker (1958) stock recruitment model to two distinct stock recruitment time-series data sets, blue line full data set and red line 1978-2012 data set (top panel), and the fit to the assessment preferred model (model 18_1), lower panel).")

fig.14_cap = figure_nums(name="fig_14",caption = "Posterior distributions of Fmsy for the two models considered in the stock productivity analysis.")

fig.15_cap = figure_nums(name="fig_15",caption = "Estimated fishery selectivity by age and year for males (panel a.) and females (panel b.).")

fig.16_cap = figure_nums(name="fig_16",caption = "Model fit to the survey biomass estimates (top left panel), model estimate of the full selection fishing mortality rate throughout the time-series (top right panel), model estimate of total biomass (middle left panel), the model estimate of survey selectivity (middle right panel) and the estimate of female spawning biomass (bottom left panel).
                         ")

fig.17_cap = figure_nums(name="fig_17",caption = "Stock  assessment model fit to the time-series of fishery and survey age composition, by sex.")

fig.18_cap = figure_nums(name="fig_18",caption = "Comparison of the fit to the survey biomass using a fixed q and the q-bottom temperature relationship.")

fig.19_cap = figure_nums(name="fig_19",caption = "Year class strength of age 5 Yellowfin Sole estimated by the stock assessment model. The dotted line is the average of the estimates from 62 years of recruitment.")

fig.20_cap = figure_nums(name="fig_20",caption = "Posterior distributions of some important parameters estimated by the preferred stock assessment model (from mcmc integration).")

fig.21_cap = figure_nums(name="fig_21",caption = "Retrospective plot of Yellowfin Sole female spawning biomass estimates (1,000s t), 2008-2016, from the recommended assessment model.")

fig.22_cap = figure_nums(name="fig_22",caption = "Projection of Yellowfin Sole female spawning biomass (1,000s t) at the average full- selection F from the past 5 years (0.104) through 2032 with B40% and Bmsy levels indicated (top panel).  Tier 1 projection of Yellowfin Sole ABC assuming 150,000 t catch in each future year for models 14_1 (base model) and 18_1 (new base model) (bottom panel).")

fig.23_cap = figure_nums(name="fig_23",caption = "Phase plane figure of the time-series of Yellowfin Sole female spawning biomass relative to the harvest control rule with 1975 and 2019 indicated.")

```

\pagebreak
# Tables

`r table_nums(name = "catch", caption = paste("Foreign and domestic datch (t) of Yellowfin Sole 1954-",thisyr,".  Foreign catches are designated as joint venture processing (JVP) and domestic annual processing (DAP). Catch for ",thisyr," was downloaded October 28, 2019.",sep=""))`
```{r catch,message=FALSE,echo=FALSE}
catchZ=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/catch_YFS.csv",header=TRUE,skip=1)
colnames(catchZ)=c("Year","Foreign","JVP","DAP","Total")
kable(catchZ,"latex",booktabs=T,linesep="",longtable=TRUE)%>%
  add_header_above(c(" ", " ","Domestic" = 2, " " ))
```

\pagebreak
`r table_nums(name = "retdis", caption = paste("Estimates of retained and discarded (t) Yellowfin Sole caught in Bering Sea fisheries through September 25th, ",thisyr,", and the proportion discarded.",sep=""))`
```{r retdis,message=FALSE,echo=FALSE}
retdis=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/RetDis_full.csv",header=TRUE)#Sept 25 2019

Year=names(table(retdis$Year))
rdtable_full=matrix(0,length(Year),3)
colnames(rdtable_full)=c("Year","Retained (mt)","Discarded (mt)")
rdtable_full[,1]=Year
for (i in 1:length(Year)){
 rdtable_full[i,2]=round(sum(retdis$Retained..mt.[which(retdis$Year==Year[i])]))
 rdtable_full[i,3]=round(sum(retdis$Discards..mt.[which(retdis$Year==Year[i])]))
}
ret=formatC(as.numeric(as.character(rdtable_full[,2])),format="d",big.mark=",")
dis=formatC(as.numeric(as.character(rdtable_full[,3])),format="d",big.mark=",")
prop=round(as.numeric(as.character(rdtable_full[,3]))/(as.numeric(as.character(rdtable_full[,2]))+as.numeric(as.character(rdtable_full[,3]))),2)

rdfull=cbind(Year,ret,dis,prop)
colnames(rdfull)=c("Year","Retained (mt)","Discarded (mt)","Proportion discarded")

kable(rdfull,"latex",booktabs=T,linesep="",longtable=TRUE,align='r')
```

\pagebreak
`r table_nums(name = "retdiscat", caption = paste("Discarded and retained catch of non-CDQ Yellowfin Sole, by fishery, in ",thisyr-1,". Gear types include longline (HAL), bottom trawl (NPT), pot (POT), and pelagic trawl (PTR). Catch was non-zero for all target-gear combinations shown, but may appear as zero as results were rounded to the nearest metric ton (mt). Source: NMFS AKRO BLEND/Catch Accounting System.",sep=""))`
```{r retdiscat,message=FALSE, echo=FALSE}
#note: get FMP groundfish total discards and discard rate by regulatory area in BSAI
#retdis2018=read.csv(here("/doc/Rmd/retdis2018.csv"),header=TRUE)
retdis2018=retdis[which(retdis$Year=="2018"),]
target=names(table(retdis2018$Target))[2:18]
gear=c("HAL","NPT","POT","PTR")
rdtab=matrix(0,length(target)*4,3)
rownames(rdtab)=rep(target,4)
colnames(rdtab)=c("Gear","Discarded","Retained")
rdtab[,1]=rep(gear,each=length(target))
for(i in 1:length(target)){
 for(j in 1:length(gear)){
 rdtab[i+((j-1)*13),2]=sum(retdis2018$Discards..mt.[which(retdis2018$Target==target[i]&retdis2018$Gear==gear[j])])
 rdtab[i+((j-1)*13),3]=sum(retdis2018$Retained..mt.[which(retdis2018$Target==target[i]&retdis2018$Gear==gear[j])])
}}
rd=data.frame(rdtab[which((as.numeric(rdtab[,2])+as.numeric(rdtab[,3]))>0),])
rd1=cbind(as.character(rownames(rd)),as.character(rd[,1]),formatC(round(as.numeric(as.character(rd[,2]))),big.mark=",",format="d"),formatC(as.numeric(as.character(rd[,3])),big.mark=",",format="d"))
colnames(rd1)=c("Trip target name","Gear type","Discarded (mt)","Retained (mt)")
rownames(rd1)=c("Greenland Turbot","Other Species","Pacific Cod","Alaska Plaice","Atka Mackerel",
"Kamchatka Flounder","No retained catch","Other Species","Pacific Cod","Pollock - bottom",
"Pollock - midwater","Atka Mackerel","Yellowfin Sole","Alaska Plaice")
kable(rd1[,2:4],"latex",booktabs=T,linesep="",longtable=FALSE)
```


\pagebreak
`r table_nums(name = "catageprop", caption =paste("Yellowfin Sole fishery catch-at-age (proportions), 1975-",thisyr-1," (female), ages 7-17+.",sep=""))`
```{r catageprop,message=FALSE,echo=FALSE}
#YFS2$catage_f from 1954-thisyr, want from 1975, ages 7 - 17+
#select 1975 to last year

a=data.frame(seq(1975,thisyr-1,1),round(femcat,4)[22:(nrow(femcat)-1),7:17],rowSums(round(femcat,4)[22:(nrow(femcat)-1),7:17]))
colnames(a)=c("Year",seq(7,16,1),"17+","Total proportion over age 7")
kable(a,"latex",booktabs=T,linesep="")%>% 
  kable_styling(latex_options="scale_down")

```





\pagebreak
`r table_nums(name = "tab_5", caption = "Yellowfin Sole biomass estimates (t) from the annual Bering Sea shelf bottom trawl survey, with upper and lower 95% confidence intervals.")`
```{r tab_5, message=FALSE,echo=FALSE}
srvbio=cbind(YFS2$yr_bts,formatC(1000*YFS2$ob_bts,format="d",big.mark=","),formatC(1000*YFS2$ob_bts-1.96*YFS2$sd_ob_bts,format="d",big.mark=","),formatC(1000*YFS2$ob_bts+1.96*YFS2$sd_ob_bts,format="d",big.mark=","))
colnames(srvbio)=c("Year","Biomass (t)","Lower confidence interval","Upper confidence interval")
kable(srvbio,"latex",booktabs=T,linesep="")


```

`r table_nums(name = "NBS_bio", caption = "Yellowfin Sole biomass estimates (t) from the northern Bering Sea survey, with upper and lower 95% confidence intervals.")`
```{r NBS_bio, message=FALSE,echo=FALSE}
NBSbiomass2019=520028.93
NBSvariance2019=3868517857.3
NBS=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/NBS_survey/NBS_YFS.csv",header=TRUE)
colnames(NBS)=c("Biomass (t)","LCI","HCI","Haul count","Hauls with catch","Number count","Length count")
kable(NBS,"latex",booktabs=T,linesep="")
```
 

\pagebreak
`r table_nums(name = "tab_6", caption = paste("Yellowfin Sole population numbers-at-age (millions) estimated from the annual EBS bottom trawl surveys, 1982-",thisyr,". Numbers-at-age from 1982-1986 come from the standard survey area, and years 1987 forward come from the 'plusnw' extended survey area. Females are presented first, followed by males.",sep=""))`
```{r tab_6}
catage_p=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/YFS_EBS_survey/yfs_agecomp_ebs_plusnw.csv",header=TRUE)
catage_s=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/YFS_EBS_survey/yfs_agecomp_ebs_standard.csv",header=TRUE)

age=seq(2,17,1)
yrs=names(table(catage_p$YEAR))

catage_f=matrix(0,length(yrs),length(age))
catage_m=matrix(0,length(yrs),length(age))

isEmpty <- function(x) 
 {
 return(identical(x, numeric(0)))
 }

for(i in 1:length(yrs)){
 for(j in 1:16)
 {
  if (yrs[i] %in% c(1982:1986)){
   if(!isEmpty(catage_s$AGEPOP[which(catage_s$YEAR==yrs[i]&catage_s$SEX==2&catage_s$AGE==(j+1)&catage_s$SUBAREA==999999)]))
   {catage_f[i,j]=catage_s$AGEPOP[which(catage_s$YEAR==yrs[i]&catage_s$SEX==2&catage_s$AGE==(j+1)&catage_s$SUBAREA==999999)]}
      if(!isEmpty(catage_s$AGEPOP[which(catage_s$YEAR==yrs[i]&catage_s$SEX==1&catage_s$AGE==(j+1)&catage_s$SUBAREA==999999)]))
    {catage_m[i,j]=catage_s$AGEPOP[which(catage_s$YEAR==yrs[i]&catage_s$SEX==1&catage_s$AGE==(j+1)&catage_s$SUBAREA==999999)]}
  }
  if (yrs[i] >=1987){
   if(!isEmpty(catage_p$AGEPOP[which(catage_p$YEAR==yrs[i]&catage_p$SEX==2&catage_p$AGE==(j+1)&catage_p$SUBAREA==999999)]))
  catage_f[i,j]=catage_p$AGEPOP[which(catage_p$YEAR==yrs[i]&catage_p$SEX==2&catage_p$AGE==(j+1)&catage_p$SUBAREA==999999)]
      if(!isEmpty(catage_p$AGEPOP[which(catage_p$YEAR==yrs[i]&catage_p$SEX==1&catage_p$AGE==(j+1)&catage_p$SUBAREA==999999)]))
  catage_m[i,j]=catage_p$AGEPOP[which(catage_p$YEAR==yrs[i]&catage_p$SEX==1&catage_p$AGE==(j+1)&catage_p$SUBAREA==999999)]
  }
 }
}

rownames(catage_f)=yrs
colnames(catage_f)=c(seq(2,16,1),"17+")

print(kable(formatC(catage_f/1000000,format="d",big.mark=","),"latex",booktabs=T,linesep="",align='r')%>%kable_styling(latex_options="scale_down")%>%add_header_above(c("Year", "Age (Females)"=16 )))

rownames(catage_m)=yrs
colnames(catage_m)=c(seq(2,16,1),"17+")

print(kable(formatC(catage_m/1000000,format="d",big.mark=","),"latex",booktabs=T,linesep="",align='r')%>%kable_styling(latex_options="scale_down")%>%add_header_above(c("Year", "Age (Males)"=16 )))

```


\pagebreak 


`r table_nums(name="liketab2",caption="Comparison of likelihood values for survey and fishery age, selectivity, survey biomass, recruitment, and total likelihood for Models 18.1 and 18.2.")`
```{r liketab, echo=FALSE}
totallike2=sum(YFS2$wt_like,YFS2$age_likeihood_for_survey,YFS2$age_likelihood_for_fishery,YFS2$catch_likelihood,YFS2$selectivity_likelihood,YFS2$survey_likelihood,YFS2$recruitment_likelilhood) #recommended
totallike1=sum(YFS1$wt_like,YFS1$age_likeihood_for_survey,YFS1$age_likelihood_for_fishery,YFS1$catch_likelihood,YFS1$selectivity_likelihood,YFS1$survey_likelihood,YFS1$recruitment_likelilhood)  
```

\begin{table}[h]
\centering
\begin{tabular}{ lrr }
\hline
 Likelihood component & Model $18.1$ & Model $18.2$ \\
\hline
Survey age	        &       `r round(YFS1$age_likeihood_for_survey,2)`  &  `r round(YFS2$age_likeihood_for_survey,2)`   \\
Fishery age  	  &       `r round(YFS1$age_likelihood_for_fishery,2)` &  `r round(YFS2$age_likelihood_for_fishery,2)`  \\
Selectivity     &       `r round(sum(YFS1$selectivity_likelihood),2)`& `r round(sum(YFS2$selectivity_likelihood),2)`\\
Survey biomass     &       `r round(YFS1$survey_likelihood,2)`& `r round(YFS2$survey_likelihood,2)`\\
Recruitment     &       `r round(sum(YFS1$recruitment_likelilhood),2)`& `r round(sum(YFS2$recruitment_likelilhood),2)`  \\
\hline
Total              &       `r round(totallike1,2)`      &`r round(totallike2,2)`      \\ 
\hline
\end{tabular}
\end{table}


\pagebreak

`r table_nums(name="Mcomparison",caption="Comparison of reference points using Model 18.1 and Model 18.2.")`
```{r Mcomparison}
```

\begin{table}[ht]
\centering
\begin{tabular}{lrr|rr}
  \hline
       & \multicolumn{2}{c|}{Model 18.2}  & \multicolumn{2}{c}{Model 18.1}               \\
        Quantity & `r thisyr+1`      &`r thisyr+2`   & `r thisyr+1`      &`r thisyr+2` \\ 
  \hline
$M$ (natural mortality rate)          &  0.12   &  0.12   &  0.12, 0.135   & 0.12, 0.135   \\
Tier                                  &  1a   &  1a   &  1a   & 1a   \\
Projected total (age 6+) biomass (t)  &  `r formatC(extra_sd2$GM_Biom[1]*1000,format="d",big.mark=",")`  &  `r formatC(extra_sd2$GM_Biom[2]*1000,format="d",big.mark=",")` t &  `r formatC(extra_sd1$GM_Biom[1]*1000,format="d",big.mark=",")` t & `r formatC(extra_sd1$GM_Biom[2]*1000,format="d",big.mark=",")` t \\
Projected female spawning biomass (t) &  `r formatC(extra_sd2$SSB[1]*1000,format="d",big.mark=",")` t &  `r formatC(extra_sd2$SSB[2]*1000,format="d",big.mark=",")`t  &  `r formatC(extra_sd1$SSB[1]*1000,format="d",big.mark=",")` t & `r formatC(extra_sd1$SSB[2]*1000,format="d",big.mark=",")` t \\
$\:\:\:\:\:\:B_{100\%}$                           &  `r B0_2`  &  `r B0_2`  &  `r B0_1` t & `r B0_1` t \\
$\:\:\:\:\:\:B_{MSY\%}$                            &  `r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd1$Bmsy[1],format="d",big.mark=",")` t & `r formatC(1000*extra_sd1$Bmsy[1],format="d",big.mark=",")` t \\
$F_{OFL}$                             &  `r round(extra_sd2$AM_Fmsyr[1],3)`   &  `r round(extra_sd2$AM_Fmsyr[2],3)`   &  `r round(extra_sd1$AM_Fmsyr[1],3)`  & `r round(extra_sd1$AM_Fmsyr[2],3)`   \\
$maxF_{ABC}$                          &  `r round(extra_sd2$HM_Fmsyr[1],3)`   &   `r round(extra_sd2$HM_Fmsyr[2],3)`   &  `r round(extra_sd1$HM_Fmsyr[1],3)`    & `r round(extra_sd1$HM_Fmsyr[2],3)`    \\
$F_{ABC}$                             &  `r round(extra_sd2$HM_Fmsyr[1],3)`   &  `r round(extra_sd2$HM_Fmsyr[2],3)`   &  `r round(extra_sd1$HM_Fmsyr[1],3)`   & `r round(extra_sd1$HM_Fmsyr[2],3)`   \\
$OFL$                                 &  `r formatC(1000*extra_sd2$OFL_AM[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd2$OFL_AM[2],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd1$OFL_AM[1],format="d",big.mark=",")` t & `r formatC(1000*extra_sd1$OFL_AM[2],format="d",big.mark=",")`t \\
$maxABC$                              &  `r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") ` &  `r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",") `  &  `r formatC(1000*extra_sd1$ABC_HM[1],format="d",big.mark=",") ` t & `r formatC(1000*extra_sd1$ABC_HM[2],format="d",big.mark=",") ` t \\
$ABC$                                 &  `r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") `  &  `r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",") ` & `r formatC(1000*extra_sd1$ABC_HM[1],format="d",big.mark=",") ` t & `r formatC(1000*extra_sd1$ABC_HM[2],format="d",big.mark=",") ` t \\
\hline
Status                               & `r thisyr-1`       & `r thisyr`      &  `r thisyr-1`             & `r thisyr`          \\
\hline
Overfishing                           & No         & n/a       & No                        & n/a                 \\
Overfished                           & n/a         & No       & n/a                        & No                  \\
Approaching overfished               & n/a         & No       & n/a                        & No                  \\
\hline
\end{tabular}
\begin{tablenotes}
\item Projections were based on xxx.
  \end{tablenotes}
\end{table}

\pagebreak

`r table_nums(name="Mcomparison",caption="Comparison of reference points with Model 18.1 and Model 18.2.")`

\begin{table}[ht]
\centering
\begin{tabular}{lrr|rr}
  \hline
       & \multicolumn{2}{c|}{Model 18.2}  & \multicolumn{2}{c}{Model 18.1}               \\
        Quantity & `r thisyr+1`      &`r thisyr+2`   & `r thisyr+1`      &`r thisyr+2` \\ 
  \hline
$M$ (natural mortality rate)          &  0.12   &  0.12   &  0.12, 0.135   & 0.12, 0.135   \\
Tier                                  &  1a   &  1a   &  1a   & 1a   \\
Projected total (age 6+) biomass (t)  &  `r formatC(extra_sd2$GM_Biom[1]*1000,format="d",big.mark=",")` t  &  `r formatC(extra_sd2$GM_Biom[2]*1000,format="d",big.mark=",")` t &  `r formatC(extra_sd1$GM_Biom[1]*1000,format="d",big.mark=",")` t & `r formatC(extra_sd1$GM_Biom[2]*1000,format="d",big.mark=",")` t \\
Projected female spawning biomass (t) &  `r formatC(extra_sd2$SSB[1]*1000,format="d",big.mark=",")` t &  `r formatC(extra_sd2$SSB[2]*1000,format="d",big.mark=",")`t  &  `r formatC(extra_sd1$SSB[1]*1000,format="d",big.mark=",")` t & `r formatC(extra_sd1$SSB[2]*1000,format="d",big.mark=",")` t \\
$\:\:\:\:\:\:B_{100\%}$                           &  `r B0_2`  &  `r B0_2`  &  `r B0_1` t & `r B0_1` t \\
$\:\:\:\:\:\:B_{MSY\%}$                            &  `r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd2$Bmsy[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd1$Bmsy[1],format="d",big.mark=",")` t & `r formatC(1000*extra_sd1$Bmsy[1],format="d",big.mark=",")` t \\
$F_{OFL}$                             &  `r round(extra_sd2$AM_Fmsyr[1],3)`   &  `r round(extra_sd2$AM_Fmsyr[2],3)`   &  `r round(extra_sd1$AM_Fmsyr[1],3)`  & `r round(extra_sd1$AM_Fmsyr[2],3)`   \\
$maxF_{ABC}$                          &  `r round(extra_sd2$HM_Fmsyr[1],3)`   &   `r round(extra_sd2$HM_Fmsyr[2],3)`   &  `r round(extra_sd1$HM_Fmsyr[1],3)`    & `r round(extra_sd1$HM_Fmsyr[2],3)`    \\
$F_{ABC}$                             &  `r round(extra_sd2$HM_Fmsyr[1],3)`   &  `r round(extra_sd2$HM_Fmsyr[2],3)`   &  `r round(extra_sd1$HM_Fmsyr[1],3)`   & `r round(extra_sd1$HM_Fmsyr[2],3)`   \\
$OFL$                                 &  `r formatC(1000*extra_sd2$OFL_AM[1],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd2$OFL_AM[2],format="d",big.mark=",")`  &  `r formatC(1000*extra_sd1$OFL_AM[1],format="d",big.mark=",")` t & `r formatC(1000*extra_sd1$OFL_AM[2],format="d",big.mark=",")`t \\
$maxABC$                              &  `r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") ` &  `r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",") `  &  `r formatC(1000*extra_sd1$ABC_HM[1],format="d",big.mark=",") ` t & `r formatC(1000*extra_sd1$ABC_HM[2],format="d",big.mark=",") ` t \\
$ABC$                                 &  `r formatC(1000*extra_sd2$ABC_HM[1],format="d",big.mark=",") `  &  `r formatC(1000*extra_sd2$ABC_HM[2],format="d",big.mark=",") ` & `r formatC(1000*extra_sd1$ABC_HM[1],format="d",big.mark=",") ` t & `r formatC(1000*extra_sd1$ABC_HM[2],format="d",big.mark=",") ` t \\
\hline
Status                               & `r thisyr-1`       & `r thisyr`      &  `r thisyr-1`             & `r thisyr`          \\
\hline
Overfishing                           & No         & n/a       & No                        & n/a                 \\
Overfished                           & n/a         & No       & n/a                        & No                  \\
Approaching overfished               & n/a         & No       & n/a                        & No                  \\
\hline
\end{tabular}
\begin{tablenotes}
\item Projections were based on an assumed catch equal to that of `r thisyr-1`.
  \end{tablenotes}
\end{table}
\pagebreak
`r table_nums(name = "tab_7", caption = "Occurrence of Yellowfin Sole in the Bering Sea trawl survey and collections of length and age structures and the number of otoliths aged from each survey.")`
```{r tab_7}
haul_oto=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/Hauls_otoliths.csv",header=TRUE)
colnames(haul_oto)=c("Year","Total hauls","Hauls with length","Number of lengths","Hauls with otoliths","Hauls with ages","Number of otoliths","Number of ages")
kable(haul_oto,"latex",booktabs=T,linesep="")%>%kable_styling(latex_options="scale_down")
```

\pagebreak
`r table_nums(name = "tab_8", caption = paste("Total tonnage of Yellowfin Sole caught in resource assessment surveys in the eastern Bering Sea from 1977-",thisyr,"."))`
```{r tab_8, message=FALSE,echo=FALSE}
rescatch=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/ResearchCatch.csv",header=TRUE)
colnames(rescatch)=c("Year","Research catch (t)")
kable(rescatch,"latex",booktabs=T,linesep="")
```

\pagebreak
`r table_nums(name = "tab_9", caption = "Mean length and weight at age for Yellowfin Sole (unsmoothed), females presented first, followed by males.")`
```{r tab_9}
thisyr=2019
nyrs=length(seq(1964,thisyr,1))
wt_f=YFS2$wt_srv_f[11:(nyrs+10),]
wt_m=YFS2$wt_srv_m[11:(nyrs+10),]
colnames(wt_f)=seq(1,20,1)
rownames(wt_f)=seq(1964,thisyr,1)
colnames(wt_m)=seq(1,20,1)
rownames(wt_m)=seq(1964,thisyr,1)
print(kable(wt_f,"latex",booktabs=T,linesep="")%>%kable_styling(latex_options="scale_down")%>%add_header_above(c("Year", "Age (Females)"=19 )))

print(kable(wt_m,"latex",booktabs=T,linesep="")%>%kable_styling(latex_options="scale_down")%>%add_header_above(c("Year", "Age (Males)"=19 )))
```


\pagebreak
`r table_nums(name = "tab_10", caption = "Female Yellowfin Sole proportion mature at age from Nichol (1995) and TenBrink and Wilderbuer (2015).")`
```{r tab_10,message=FALSE,echo=FALSE}
mat=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/maturity_T10.csv",header=TRUE)
colnames(mat)=c("Age","1992, 1993 samples","2012 samples","Combined")
kable(mat,"latex",booktabs=T,linesep="")

```

\pagebreak
`r table_nums('tab_11')`
```{r tab_11,message=FALSE,echo=FALSE}



```



\pagebreak
# Figures


```{r NBS,echo=FALSE,message=FALSE}
```
\begin{figure}[h]
    \centering
    \includegraphics[width=0.495\textwidth]{/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/NBS_survey/yfs10bw.png}
    \includegraphics[width=0.495\textwidth]{/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/NBS_survey/yfs17bw.png}
    \includegraphics[width=0.495\textwidth]{/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/NBS_survey/yfs18bw.png}
    \includegraphics[width=0.495\textwidth]{/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/NBS_survey/yfs19bw.png}
\end{figure}

`r figure_nums(name = "NBS", caption ="Distribution of Yellowfin Sole in the eastern and northern Bering sea based on surveys conducted in 2010 (upper left), 2017 (upper right), 2018 (lower left), and 2019 (lower right).")`

    
    
```{r fig_1, message=FALSE, echo=FALSE,fig.width = 5.1, fig.height = 4}  
catchtab=data.frame(cbind(YFS2$Yr,YFS2$Obs_catch))
colnames(catchtab)=c("Year","Catch")
ggplot(data=catchtab,aes(x=Year,y=Catch))+geom_line(aes(x=Year,y=Catch))+
theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Catch (x 1,000 t)")+scale_x_continuous(breaks = seq(1954,2018, by = 2))+theme(axis.text.x=element_text(angle=90,hjust=1))+
 theme(legend.position="right")+scale_color_discrete(name="")
```

```{r fig_1b, message=FALSE, echo=FALSE,fig.width = 6.25, fig.height = 4}
#Figure 1 lower panel
#blend_catchYFS=read.csv("/Users/ingridspies/YFS_Blend_CAS.csv",header=TRUE)
#yrs=c(2003:2018)
#cat_month=matrix(0,length(yrs),12)
#cat_month2=matrix(0,length(yrs),12)
#for (i in 1:length(yrs)){
#cat_month[i,]=table(blend_catchYFS2$Month[which(blend_catchYFS2$Year==yrs[i])])
#}
#colnames(cat_month)=c(1:12)
#rownames(cat_month)=yrs
#cat_month1=cat_month/rowSums(cat_month)
#for(i in 1:length(yrs)){
#cat_month2[i,]=cumsum(cat_month1[i,])
#}
#Cat=c(cat_month2)
#Year=rep(yrs,12)
#Month=rep(seq(1,12,1),each=length(yrs))
#catch_month=data.frame(Cat,Year,Month)
#write.csv(catch_month,here("/doc/Rmd/catch_month.csv"))

catch_month=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/catch_month.csv")
ggplot(data=catch_month, aes(x=Month,y=Cat,group=factor(Year)))+geom_line(aes(colour=Year),size=1.2)+scale_x_continuous(breaks = seq(1,12, by = 1))+theme(legend.position="right")+labs(y = "Relative cumulative catch")+scale_colour_gradient(low="lightblue",high="darkblue")+theme_few(base_size=14)

```
\
`r figure_nums('fig_1')`

\pagebreak

```{r fig_3,fig.align='center',message=FALSE, echo=FALSE,fig.align='center'}  
library(scales)
par(mar=c(5,6,6,1))
area=names(table(retdis2018$Reporting.Area.Code))
areaz=vector()
for(i in 1:length(area)){
 areaz[i]=sum(retdis2018$Total.Catch..mt.[which(retdis2018$Reporting.Area.Code==area[i])])
}
area_res=as.data.frame(cbind(area,round(areaz)))
area_res2=area_res[which(as.numeric(as.character(area_res[,2]))>1000),]
area_res3=data.frame(as.factor(as.character(area_res2[,1])),as.numeric(as.character(area_res2[,2])))
area_res4=cbind(area_res3,area_res3[,2]/sum(area_res3[,2]))
colnames(area_res4)=c("NMFS.Area","Catch_full","Catch")
area_res5=cbind(area_res4,cumsum(area_res4$Catch)-(area_res4$Catch/2))
colnames(area_res5)=c("NMFS.Area","Catch_full","Catch","pos")

blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

ggplot(area_res5, aes(x="", y=Catch, fill=NMFS.Area))+
geom_bar(width = 2, stat = "identity")+coord_polar("y", start=0,direction=1)+blank_theme+theme(axis.text.x=element_blank()) +geom_text(aes(label = NMFS.Area),position = position_stack(vjust = 0.5), size=c(8,4,4,4,8,8,8))+ scale_fill_discrete(name = "NMFS Area")+ theme(legend.text=element_text(size=15),legend.title=element_text(size=16))

```
\
 
```{r fig_3b, message=FALSE, echo=FALSE,fig.align='center'} 
#blend_catch2018=blend_catchYFS[which(blend_catchYFS2$Year==2018),]
#cat_mo2018=data.frame(cbind(c("January","February","March","April","May","June","July","August","September","October","November","December"),table(blend_catch2018$Month)))
#cat_mo2018b=cbind(as.character(cat_mo2018[,1]),formatC(as.numeric(as.character(cat_mo2018[,2])),format="d",big.mark=","))
#write.csv(cat_mo2018b,here("/doc/Rmd/cat_mo2018b.csv"))
cat_mo2018b=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/cat_mo2018b.csv")
colnames(cat_mo2018b)=c("N","Month","Catch")
cat_mo2018b$Month=factor(cat_mo2018b$Month,levels=cat_mo2018b$Month)
cat_mo2018b$Catch=as.numeric(as.character(cat_mo2018b$Catch))
Proportion=percent(round(cat_mo2018b$Catch/sum(cat_mo2018b$Catch),2))
cm=cbind(cat_mo2018b,Proportion)

ggplot(cm, aes(x="", y=Catch, fill=Month))+
geom_bar(width = 2, stat = "identity")+coord_polar("y", start=0,direction=-1)+blank_theme+theme(axis.text.x=element_blank()) +geom_text(aes(label = Proportion),position = position_stack(vjust =0.5),vjust=c(-.3,rep(1.1,11)),hjust=c(rep(.5,11),.3), size=c(5,rep(5,11)))+ scale_fill_discrete(name = "Month")+ theme(legend.text=element_text(size=15),legend.title=element_text(size=16))
 
```
\
\
`r figure_nums('fig_3')`

\pagebreak

```{r fig_6, message=FALSE, echo=FALSE}  
LCI=YFS2$ob_bts-1.96*YFS2$sd_ob_bts
HCI=YFS2$ob_bts+1.96*YFS2$sd_ob_bts
srv_bio=data.frame(cbind(YFS2$yrs_srv,YFS2$ob_bts,YFS2$sd_ob_bts,LCI,HCI))

colnames(srv_bio)=c("Year","Biomass","sd","LCI","HCI")
ggplot(data=srv_bio,aes(x=Year,y=Biomass))+geom_point(aes(x=Year,y=Biomass))+geom_errorbar(aes(ymin=LCI, ymax=HCI), width=.1)+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Biomass (x 1,000 t)")+scale_x_continuous(breaks = seq(1982,2019, by = 2))+theme(axis.text.x=element_text(angle=90,hjust=1))+ylim(0,4500)+
 theme(legend.position="right")+scale_color_discrete(name="")

```
`r figure_nums('fig_6')`

\pagebreak

```{r fig_8, message=FALSE, echo=FALSE}  
wts2=data.frame(cbind(seq(1,20,1)),colMeans(YFS2$wt_srv_f),colMeans(YFS2$wt_srv_m))
colnames(wts2)=c("Age","Females","Males")
ggplot(data=wts2)+geom_line(aes(x=Age,y=Females,colour="Females"))+geom_line(aes(x=Age,y=Males,colour="Males"))+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Weight (g)")+
 theme(legend.position="right")+scale_color_discrete(name="Sex")

```
`r figure_nums('fig_8')`

\pagebreak

```{r fig_9, message=FALSE, echo=FALSE}  

```

\begin{center}
\includegraphics[width=5in,angle=270,origin=c]{/Users/ingridspies/admbmodels/flatfish/assessments/yfs/doc/data/Black_Fig3.pdf}
\end{center}
`r figure_nums('fig_9')`


\pagebreak

```{r fig_10, message=FALSE, echo=FALSE}  
env=data.frame(cbind(YFS2$yrs_srv,YFS2$Bottom_temp[1,]))
colnames(env)=c("Year","Temp")
ggplot(data=env)+geom_line(aes(x=Year,y=Temp))+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+ylab("Temperature Anomaly")+theme(legend.position="right")+scale_color_discrete(name="")+scale_x_continuous(breaks = seq(1982,2018, by = 2))+theme(axis.text.x=element_text(angle=90,hjust=1))+geom_hline(yintercept=0,linetype=1)

```
`r figure_nums('fig_10')`

\pagebreak

```{r fig_16, message=FALSE, echo=FALSE}  
#fit to survey biomass
LCI=YFS2$ob_bts-1.96*YFS2$sd_ob_bts
HCI=YFS2$ob_bts+1.96*YFS2$sd_ob_bts
srv_bio2=data.frame(cbind(YFS2$yrs_srv,YFS2$ob_bts,YFS2$sd_ob_bts,LCI,HCI,YFS2$pred_srv[29:65]))
colnames(srv_bio2)=c("Year","Biomass","sd","LCI","HCI","Predicted")
ggplot(data=srv_bio2,aes(x=Year,y=Biomass))+geom_point(aes(x=Year,y=Biomass,col="Survey Estimate"))+geom_errorbar(aes(ymin=LCI, ymax=HCI), width=.1)+geom_line(aes(x=Year,y=Predicted,col="Model Estimate"))+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Biomass (x 1,000 t)")+scale_x_continuous(breaks = seq(1982,2018, by = 2))+theme(axis.text.x=element_text(angle=90,hjust=1))+ylim(0,4500)+
 theme(legend.position="right")+scale_color_discrete(name="")

#total biomass
TotBiom=data.frame(YFS2$TotBiom)
colnames(TotBiom)=c("Year","Biomass","sd","LCI","HCI")
ggplot(data=TotBiom,aes(x=Year,y=Biomass))+geom_errorbar(aes(ymin=LCI, ymax=HCI), width=.1)+geom_point(aes(x=Year,y=Biomass))+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Biomass (x 1,000 t)")+scale_x_continuous(breaks = seq(1954,2018, by = 4))+theme(axis.text.x=element_text(angle=90,hjust=1))+ylim(0,4500)

#Female spawning biomass (SSB seems like it)
SSB=data.frame(YFS2$SSB)
colnames(SSB)=c("Year","FSB","sd","LCI","HCI")
ggplot(data=SSB,aes(x=Year,y=FSB))+geom_errorbar(aes(ymin=LCI, ymax=HCI), width=.1)+geom_point(aes(x=Year,y=FSB))+theme_bw()+theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Biomass (x 1,000 t)")+scale_x_continuous(breaks = seq(1954,2018, by = 4))+theme(axis.text.x=element_text(angle=90,hjust=1))+ylim(0,1500)

#survey selectivity
srvsel=data.frame(cbind(seq(1,20,1),YFS2$sel_srv_f,YFS2$sel_srv_m))
colnames(srvsel)=c("Age","Female","Male")
ggplot(data=srvsel)+geom_line(aes(x=Age,y=Female,col="Female"))+geom_line(aes(x=Age,y=Male,col="Male"))+theme_bw()+
 theme(legend.title=element_text(size=13),legend.text=element_text(size=11),axis.text=element_text(size=11),axis.title.x = element_text(size=13),axis.title.y = element_text(size=14))+
 ylab("Proportion selected")+scale_x_continuous(breaks = seq(1,20, by = 2))+scale_color_discrete(name="")
```
`r figure_nums('fig_16')`


Nichol, D.G., Kotwicki, S., Wilderbuer, T.K., Lauth, R.R. and Ianelli, J.N., 2019. Availability of yellowfin sole Limanda aspera to the eastern Bering Sea trawl survey and its effect on estimates of survey biomass. Fisheries Research, 211, pp.319-330.