---
title: "Eastern Bering Sea bottom temperature and growth of Yellowfin Sole"
author: "Ingrid Spies"
date: "3/4/2022"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(captioner)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(coldpool)
library(fishmethods)
library(ggnewscale)
library(car)#for leveneTest
table_nums <- captioner::captioner(prefix = "Table ",levels=1,auto_space = FALSE)
figure_nums <- captioner::captioner(prefix="Figure ",levels = 1, auto_space = FALSE)


tempz=read.csv("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/GrowthAnal/tempz.csv",header=TRUE)


source("/Users/ingridspies/admbmodels/Katch/assessments/R/read-rep.R")
YFS2020=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m2020/fm.rep")#last years model 
YFS18_2=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2/fm.rep")#this is the preferred model. Male M is estimated.

YFS18_2growth_SSP585=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2_growth/fm_SSP585.rep")#this is the preferred model with SSP585 future

YFS18_2growth_SSP126=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2_growth/fm_SSP126.rep")#this is the preferred model with SSP585 future

YFS18_2growth_mnhind=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2_growth/fm_mnhind.rep")#this is the preferred model with SSP585 future

ABCstd=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2/ABC_OFL.rep")
ABCgrowth=read.rep("/Users/ingridspies/admbmodels/flatfish/assessments/yfs/runs/m18_2_growth/ABC_OFL.rep")

colnames(ABCstd$Year)=c("Year","HM_Fmsyr","AM_Fmsyr","GM_Biom","Catch_Assump","ABC_HM","OFL_AM","Bmsy","SSB","Adjust","CV_Biom","CV_SSB","ABC_T3","OFL_T3")
colnames(ABCgrowth$Year)=c("Year","HM_Fmsyr","AM_Fmsyr","GM_Biom","Catch_Assump","ABC_HM","OFL_AM","Bmsy","SSB","Adjust","CV_Biom","CV_SSB","ABC_T3","OFL_T3")

ABCstuff=rbind(ABCstd$Year[1,],ABCgrowth$Year[1,])

rownames(ABCstuff)=c("Standard","Growth")
ABCstuff[,c(1,5,8,9,10,13,14)]


F40=  0.12
F35=  0.16
F30=  0.21

#Line 1826

#Plot total biomass SS126
futureSSP126=t(YFS18_2growth_SSP126$future_TotBiom)
colnames(futureSSP126)=rev(c("Fixed_F","F=0","meanF","F35","F40"))
rownames(futureSSP126)=c(2021:2099)
futureSSP126_2=reshape2::melt(futureSSP126)
colnames(futureSSP126_2)=c("Year","Fishing_Mortality_Rate","TotalBiomass")
futureSSP126_2$Scenario="SSP126"

#Plot total biomass SS585
futureSSP585=t(YFS18_2growth_SSP585$future_TotBiom)
colnames(futureSSP585)=rev(c("Fixed_F","F=0","meanF","F35","F40"))
rownames(futureSSP585)=c(2021:2099)
futureSSP585_2=reshape2::melt(futureSSP585)
colnames(futureSSP585_2)=c("Year","Fishing_Mortality_Rate","TotalBiomass")
futureSSP585_2$Scenario="SSP585"

#Plot total biomass mnhind
futuremnhind=t(YFS18_2growth_mnhind$future_TotBiom)
colnames(futuremnhind)=rev(c("Fixed_F","F=0","meanF","F35","F40"))
rownames(futuremnhind)=c(2021:2099)
futuremnhind_2=reshape2::melt(futuremnhind)
colnames(futuremnhind_2)=c("Year","Fishing_Mortality_Rate","TotalBiomass")
futuremnhind_2$Scenario="mnhind"

#all 3 scenarios
All_TotBiom=rbind(futureSSP126_2,futureSSP585_2,futuremnhind_2)

ggplot(data=All_TotBiom)+geom_line(aes(x=Year,y=TotalBiomass,col=Fishing_Mortality_Rate,lty=Scenario))

#try just meanF
#Figure 6
ggplot(data=All_TotBiom[which(All_TotBiom$Fishing_Mortality_Rate=="meanF"),])+geom_line(aes(x=Year,y=TotalBiomass,col=Scenario),lwd=1.05)+theme_bw()+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))+ylab("Total Biomass (t)")+ylim(0,2800)

#Plot number at age
futureSSP126=rowSums(YFS18_2growth_SSP126$nage_future_f)
futureSSP585=rowSums(YFS18_2growth_SSP585$nage_future_f)
futuremnhind=rowSums(YFS18_2growth_mnhind$nage_future_f)
n_fut=cbind(seq(2022,2099,1),futureSSP126,futureSSP585,futuremnhind)
colnames(n_fut)=c("Year","Females_126","Females585","mnhind")
n_fut_2=reshape2::melt(n_fut)[79:312,]
n_fut_2$Year=rep(seq(2022,2099),3)
colnames(n_fut_2)=c("X","Scenario","Numbers","Year")
n_fut_2=n_fut_2[,2:4]

ggplot(n_fut_2)+geom_line(aes(x=Year,y=Numbers,col=Scenario))+theme_bw()

#Try to plot the difference between 1982-2021 in the regular code and the growth code TotalBiomass
compare_TotBio=cbind(YFS18_2$TotBiom[,2],
YFS18_2growth_mnhind$TotBiom[,2])
colnames(compare_TotBio)=c("Standard","Growth")
compare_TotBio2=reshape2::melt(compare_TotBio)
compare_TotBio2$Year=rep(seq(1954,2021,1),2)
colnames(compare_TotBio2)=c("X","Scenario","TotalBiomass","Year")
ggplot(compare_TotBio2)+geom_line(aes(x=Year,y=TotalBiomass,col=Scenario))

#Figure 6a
ggplot(compare_TotBio2[which(compare_TotBio2$Year>1981),])+geom_line(aes(x=Year,y=TotalBiomass,col=Scenario),lwd=1.1)+theme_bw()+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))+ylab("Total Biomass (t)")+ylim(0,4000)

#Try to plot the difference between 1982-2021 in the regular code and the growth code numbers at age (females)
compare_natage_f=data.frame(cbind(YFS18_2$natage_f[,2],YFS18_2growth_mnhind$natage_f[,2]))
compare_natage_f$Year=seq(1954,2021,1)
colnames(compare_natage_f)=c("Standard","Growth","Year")
compare_natage_f2=reshape2::melt(compare_natage_f,id="Year")

colnames(compare_natage_f2)=c("Year","Scenario","Numbers")
ggplot(compare_natage_f2)+geom_line(aes(x=Year,y=Numbers,col=Scenario))+theme_bw()+ggtitle("Females")

colnames(compare_natage_f2)=c("Year","Scenario","Numbers")

#Figure 7
ggplot(compare_natage_f2[which(compare_natage_f2$Year>1981),])+geom_line(aes(x=Year,y=Numbers,col=Scenario),lwd=1.1)+theme_bw()+ggtitle("Females")+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))+ylab("Number of fish (billions)")+ylim(0,4)

compare_natage_m=data.frame(cbind(YFS18_2$natage_m[,2],YFS18_2growth_mnhind$natage_m[,2]))
compare_natage_m$Year=seq(1954,2021,1)
colnames(compare_natage_m)=c("Standard","Growth","Year")
compare_natage_m2=reshape2::melt(compare_natage_m,id="Year")

colnames(compare_natage_m2)=c("Year","Scenario","Numbers")
ggplot(compare_natage_m2)+geom_line(aes(x=Year,y=Numbers,col=Scenario))+theme_bw()+ggtitle("Males")

ggplot(compare_natage_m2[which(compare_natage_m2$Year>1981),])+geom_line(aes(x=Year,y=Numbers,col=Scenario))+theme_bw()+ggtitle("Males")

#compare weightat age

plot(seq(1954,2021,1),YFS18_2growth_SSP126$wt_pop_f[,8],ylim=c(0,350))#This changes
lines(seq(1954,2021,1),YFS18_2$wt_srv_f[,8])#This changes

plot(seq(1954,2021,1),YFS18_2$wt_srv_f[,9],type="l")
lines(seq(1954,2021,1),YFS18_2growth_SSP585$wt_srv_f[,9],col="red",lwd=2)

#Try the mean over 1982-2021 
#how do you do the mean? Weighted?

#Maybe try calculating the weight of the whole population by numbers*weight so total weight at age Females age 2-13 only 

Standard=rowSums(YFS18_2$wt_srv_f[29:68,2:13]*YFS18_2$natage_f[29:68,2:13])

Growth=rowSums(YFS18_2growth_SSP585$wt_srv_f[29:68,2:13]*YFS18_2growth_SSP585$natage_f[29:68,2:13])
Year=seq(1982,2021,1)
Comp_WtBio=data.frame(rep(Year,2),c(Growth,Standard),c(rep("Growth",40),rep("Standard",40)))
colnames(Comp_WtBio)[1]="Year";colnames(Comp_WtBio)[2]="Biomass";colnames(Comp_WtBio)[3]="Scenario"
ggplot(data=Comp_WtBio)+geom_line(aes(x=Year,y=Biomass,col=Scenario),lwd=1.1)+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))+ylab("Female biomass age 2-13")+theme_bw()+ylim(0,1800)

#Here try just total weight

Standard=rowSums(YFS18_2$wt_srv_f[29:68,8:13])
Growth=rowSums(YFS18_2growth_SSP585$wt_srv_f[29:68,8:13])
Year=seq(1982,2021,1)
Comp_WtBio=data.frame(rep(Year,2),c(Growth,Standard),c(rep("Growth",40),rep("Standard",40)))
colnames(Comp_WtBio)[1]="Year";colnames(Comp_WtBio)[2]="Biomass";colnames(Comp_WtBio)[3]="Scenario"
ggplot(data=Comp_WtBio)+geom_line(aes(x=Year,y=Biomass,col=Scenario),lwd=1.1)+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))+ylab("Sum of female numbers age 2-13")+theme_bw()

#in the growth model, wt_fsh_f looks good but not wt_pop_f or wt_srv_f


#which age class is the heavisest over time? Females
ageclassF=YFS18_2$wt_srv_f[29:68,2:19]*YFS18_2$natage_f[29:68,2:19]
ageclassM=YFS18_2$wt_srv_m[29:68,2:19]*YFS18_2$natage_m[29:68,2:19]
maxage_f=vector();maxage_m=vector()
for(i in 1:40)
{maxage_f[i]=which(ageclassF[i,]==max(ageclassF[i,]))+1
maxage_m[i]=which(ageclassM[i,]==max(ageclassM[i,]))+1
}

maxage2=data.frame(cbind(seq(1982,2021),maxage_f,maxage_m))
colnames(maxage2)=c("Year","Females","Males")
maxage3=reshape2::melt(maxage2,id="Year")
colnames(maxage3)=c("Year","Sex","Age")
ggplot(maxage3)+geom_point(aes(x=Year,y=Age,col=Sex),size=1.8)+theme_bw()+theme_bw()+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+ggtitle("Age with maximum biomass (no selectivity)")+theme(axis.title=element_text(size=15))

#This represents the heaviest age class by yaer for females.
#You can see that the age with the heaviest hear class has become younger since the year 2000. This actually means that the total biomass would become less over time. It might also mean that the natural mortality of YFS is increasing over time and that this is not being incororated in the model Other explanations could also be that fishing selectiivty is getting higher at higher ages. It can also mean that maybe the older fish are not moving into the survey area as much during teh survey season.



#This changes
+ )

 #Future output
#future_TotBiom
#nage_future_f

coldpool=read.csv("/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/coldpoolindex.csv")

require(dplyr)
#count(srv_wtage_1971_2019,Year)

srv_wtage=read.csv("/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/survey_wtage/YFSageSurvey1954_2021.csv",header=TRUE)

min(as.numeric(table(srv_wtage$Year)[12:49]))
max(as.numeric(table(srv_wtage$Year)[12:49]))
mean(as.numeric(table(srv_wtage$Year)[12:49]))
sqrt(var(as.numeric(table(srv_wtage$Year)[12:49])))
sum(as.numeric(table(srv_wtage$Year)[12:49]))

#The next exploration of growth is in the increments from one year to the next
# I calculated growth increments for years 1982 to 2019 and then I compared them with the temperature in the first summer
#Here are lengths



#Here I compile all the length by age data
age=seq(1,20,1)
yrs=as.numeric(names(table(srv_wtage$Year)))
mat_lenage_srv_F=matrix(0,length(yrs),20)
colnames(mat_lenage_srv_F)=seq(1,20,1)
rownames(mat_lenage_srv_F)=yrs
mat_lenage_srv_M=mat_lenage_srv_F
for (i in 1:length(yrs)){
 for(j in 1:19){
  mat_lenage_srv_F[i,j]=mean(srv_wtage$LENGTH[which(srv_wtage$Year==yrs[i]&srv_wtage$AGE==age[j]&srv_wtage$SEX=='2'&srv_wtage$REGION=="BS")],na.rm=TRUE)
  mat_lenage_srv_F[i,20]=mean(srv_wtage$LENGTH[which(srv_wtage$Year==yrs[i]&srv_wtage$AGE>19&srv_wtage$SEX=='2'&srv_wtage$REGION=="BS")],na.rm=TRUE)
  mat_lenage_srv_M[i,j]=mean(srv_wtage$LENGTH[which(srv_wtage$Year==yrs[i]&srv_wtage$AGE==age[j]&srv_wtage$SEX=='1'&srv_wtage$REGION=="BS")],na.rm=TRUE)
  mat_lenage_srv_M[i,20]=mean(srv_wtage$LENGTH[which(srv_wtage$Year==yrs[i]&srv_wtage$AGE>19&srv_wtage$SEX=='1'&srv_wtage$REGION=="BS")],na.rm=TRUE)
 }
}
#write.csv(round(mat_lenage_srv_F),"/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/mat_lenage_srv_F.csv")
#write.csv(round(mat_lenage_srv_M),"/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/mat_lenage_srv_M.csv")

mat_lenage_srv_F=read.csv("/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/mat_lenage_srv_F.csv",header=TRUE)
mat_lenage_srv_M=read.csv("/Users/ingridspies/Documents/WorkDellStuff/Assessments/YFS/2021/mat_lenage_srv_M.csv",header=TRUE)


lenF=mat_lenage_srv_F[1:49,]
lenM=mat_lenage_srv_M[1:49,]

N=c(rep(18,18),18,18,rev(seq(2,17,1)))#This represents the number of years a cohort is represented by the data  and removed 2020
Year=seq(1982,2018,1)
matM_len=matrix(0,length(Year),length(seq(2,19,1)))
matF_len=matrix(0,length(Year),length(seq(2,19,1)))
tempMAT_len=matrix(0,length(Year),length(seq(2,19,1)))
colnames(matM_len)=seq(2,19,1);colnames(matF_len)=seq(2,19,1)
rownames(matF_len)=Year;rownames(matM_len)=Year
for(j in 1:length(N)){
 for(i in 1:N[j]){
  matM_len[j,i]=lenM[(9+i+j+2),(i+3)]-lenM[(9+i+j+1),(i+2)]
  matF_len[j,i]=lenF[(9+i+j+2),(i+3)]-lenF[(9+i+j+1),(i+2)]
  tempMAT_len[j,i]=tempz[which(tempz$YEAR==1982+i+j-1),3]
 }}

colnames(tempMAT_len)=seq(2,19,1);colnames(matM_len)=seq(2,19,1);colnames(matF_len)=seq(2,19,1)
#cbind(Year,tempMAT_len);cbind(Year,matM_len);cbind(Year,matF_len)
rownames(tempMAT_len)=Year;rownames(matM_len)=Year;rownames(matF_len)=Year

matM2_len=reshape2::melt(matM_len,id=c("Year"))
matF2_len=reshape2::melt(matF_len,id=c("Year"))
tempMAT2_len=reshape2::melt(tempMAT_len,id=c("Year"))
colnames(tempMAT2_len)=c("Year","Age1","Temperature")
colnames(matM2_len)=c("Year","Age1","Growth_Male")
colnames(matF2_len)=c("Year","Age1","Growth_Female")
#convert to cm for the whole project
matM2_len$Growth_Male=matM2_len$Growth_Male/10
matF2_len$Growth_Female=matF2_len$Growth_Female/10
OUT_len=cbind(tempMAT2_len,matM2_len$Growth_Male,matF2_len$Growth_Female)
colnames(OUT_len)=c("Year","Age1","Temperature","Growth_Male","Growth_Female")

OUT_len2=OUT_len[which(OUT_len$Temperature!=0),]
#Look at the data
#Males and females grow in length more early in life
OUT_len2$Age_factor=as.factor(OUT_len2$Age1)#set up age as a factor


```

## Introduction

Year-class strength of flatfishes is thought to be determined during the first few years of life between the pelagic egg and benthic settlement (van der Veer et al., 2015). One contributing factor to early life history year-class strength is temperature. Temperature in the early life stages can affect egg size, larval duration, size at settlement, as well as the size of suitable nursery habitat (Yeung and Cooper 2019). 

Understanding how external forces affect growth in commercially targeted groundfish is key to optimizing biomass estimates and quota setting. Yellowfin Sole _Limanda aspera_, is the most valuable commercially targeted flatfish in the Eastern Bering Sea. The stock assessment for Yellowfin Sole includes growth as a parameter in calculations that provide acceptable biological catch for the Eastern Bering Sea (Spies et al. 2021). Therefore, an investigation of temperature-mediated growth as a covariate in the model is warranted. 

Temperature may affect Yellowfin Sole in multiple aspects of their life history. It has been hypothesized that colder bottom temperatures delay migration from offshore overwintering areas to inshore spawning in summer (Nichol et al. 2019). As a result, mature individuals may reside in nearshore nursery grounds during months in which the NMFS survey occurs, which likely decreases survey biomass estimates during cold years (Nichol et al., 2019; Yeung and Cooper 2019). Survey timing may therefore also affect the size of the fish that are observed.

Yellowfin Sole may be less sensitive to temperature than a closely related flatfish, Northern Rock Sole _Lepidopsetta polyxystra_, due to differences in settlement timing. YFS settle later in summer, when the influence of the cold pool is weaker and nearshore bottom temperature is relatively stable and high (Yeung and Yang, 2018). In contrast, YFS migrate across the shelf to spawn near their nursery habitat, rather than relying on currents for larval transport to nursery habitat (Nichol and Acuna, 2001); therefore, their larvae may be less susceptible to variable currents (Yeung and Cooper 2019). Yellowfin Sole remain in the shallow nearshore nursery areas through at least their first 2 years post-settlement. They begin to disperse offshore age 3-5 and by 5-8 years they follow adult migratory patterns. There is no strong evidence thus far that Yellowfin sole have shifted distribution in response to climate change (Spies et al. 2021). 

There is some evidence of temperature-dependent growth by Yellowfin sole (Yeung et al. 2021). Yellowfin Sole otolith growth increment chronologies appear to be strongly correlated with summertime eastern Bering Sea bottom temperatures (Matta et al. 2010). The $R^2$ values of Yellowfin Sole, Alaska Plaice, and Northern Rock Sole with the EBS bottom temperatures and otolith growth increments were 0.81, 0.61, and 0.34, respectively (Matta et al. 2010). 

We examined growth increments of Yellowfin Sole from 1982-2019 to determine whether there is a correlation between growth and summer ocean bottom temperatures. We applied our results to an age structured model to determine whether modeling temperature-mediated growth would improve model fit. 

# Methods  

Data consisted of `r formatC(nrow(srv_wtage[which(srv_wtage$Year>1981&srv_wtage$Year<2020&srv_wtage$LENGTH>0&srv_wtage$AGE>0),]),big.mark=",")` lengthed Yellowfin Sole with otoliths
collected from NMFS surveys between 1982 and 2019. Otoliths were aged using break and burn methodology in >90% of otoliths (Christensen et al. 1984), and alternatively surface reading if the age was obvious. Age interpretation was performed with two readers for quality control and precision-testing. Surveys were conducted during summer months: between May through September, with start dates from May 23 through June 11 and end dates from July 12 - September 11. Data was not incorporated from 2020 because there was no survey in that year.

In addition, we examined several statistics associated with temperature for this analysis. Growth increments were associated with the AFSC interpolated bottom temperature for <100m, the depth preference for Yellowfin Sole. 

Growth increments were calculated for ages 2 to 3, 3 to 4, ..., through 19 to 20 for years starting with the 1980 cohort (which was age 2 in 1982) through the 2016 cohort (age 2 in 2018, age 3 in 2019). Growth increments were compared with bottom temperature using several statistical tests. 

First we plotted growth by age to visually assess how variability in growth changed by age `r figure_nums("agetmp",display="cite")`. Secondly, we used a Levene's test for relative variation (Schultz 1985) to examine whether there was a significant difference in variance of growth by age. We used the analysis of covariance (ANOVA) to test whether the temperature or age, or the interaction of the two, had a significant effect growth increments. 

( --rule out that length-weight relationship has changed (it has not))


# Results

Female growth was on average `r round(mean((as.vector(tapply(OUT_len2$Growth_Female,OUT_len2$Age_factor,mean,na.rm=TRUE))-as.vector(tapply(OUT_len2$Growth_Male,OUT_len2$Age_factor,mean,na.rm=TRUE)))[1:12]),1)` cm higher in females than males from age 2 to 3 through age 13 to 14 (`r figure_nums("agetmp",display="cite")`). Variability measured by the coefficient of variance (CV) remained below 1 through the growth increment starting at age `r min(as.vector(which(sqrt(tapply(OUT_len2$Growth_Female,OUT_len2$Age_factor,var,na.rm=TRUE))/tapply(OUT_len2$Growth_Female,OUT_len2$Age_factor,mean,na.rm=TRUE)>1)))` for females and through the growth increment starting at age `r min(as.vector(which(sqrt(tapply(OUT_len2$Growth_Male,OUT_len2$Age_factor,var,na.rm=TRUE))/tapply(OUT_len2$Growth_Male,OUT_len2$Age_factor,mean,na.rm=TRUE)>1)))` for males, and exceeded 1 at older age increments. Levene's test indicated that growth was significantly correlated with age in females (p = 0.043) and males (p = 0.025) (`r table_nums("levene",display="cite")`). When the ages were limited to those with growth CV < 1, growth was still significantly correlated with age for females ages 2 to 3 through 12 to 13 (p < 0.024) and males ages 2 to 3 through 11 to 12 (p = 0.020).

CV females
round(sqrt(tapply(OUT_len2$Growth_Female,OUT_len2$Age_factor,var,na.rm=TRUE))/tapply(OUT_len2$Growth_Female,OUT_len2$Age_factor,mean,na.rm=TRUE),2)

round(sqrt(tapply(OUT_len2$Growth_Male,OUT_len2$Age_factor,var,na.rm=TRUE))/tapply(OUT_len2$Growth_Male,OUT_len2$Age_factor,mean,na.rm=TRUE),2)

Analyses also considered whether a better index would be the area of the Bering Sea that was covered by ice, also known as the coldpool index. The coldpool index and the bottom temperature were very similar, so we used the Bering Sea bottom temperature (`r figure_nums("coldpooltmp",display="cite")`). The ANOVA and further analysis of how temperature may affect growth was limited to growth increments were limited to ages in which the CV was <1; for females ages 2 to 3 through 13 to 14 and males ages 2 to 3 through 12 to 13 (`r table_nums("anova",display="cite")`). The interaction between age and temperature was not significant for males or females, and therefore was omitted in the ANOVA. However, age and temperature were significant for males and females (`r table_nums("anova",display="cite")`). Age had a negative effect on growth while temperature had a positive effect (`r figure_nums("gtemp",display="cite")`).




# Discussion

Raw data from the survey and fishery are used to input 
weight at age into the model. So modelling higher growth by temperature does not seem useful. There is some limited utility in predicted future weights given the estimated future temperatures. If temperature increases to 5 for the following 5 years, then we can predict that YFS will grow larger. 

# References

Christensen, J.M., 1964. Burning of otoliths, a technique for age determination of soles and other fish. ICES Journal of Marine Science, 29(1), pp.73-81.

Essington, T.E., Matta, M.E., Black, B.A., Helser, T.E. and Spencer, P.D., 2022. Fitting growth models to otolith increments to reveal time-varying growth. Canadian Journal of Fisheries and Aquatic Sciences, 79(1), pp.159-167.

Schultz, B.B., 1985. Levene's test for relative variation. Systematic Zoology, 34(4), pp.449-456.

Spies, I., Haehn, R., Siddon, E., Conner, J., Markowitz, E., Yeung, C., Ianelli, J. 2021. Assessment of the Yellowfin Sole Stock in the Bering Sea and Aleutian Islands. In Stock Assessment and Fishery Evaluation Document for Groundfish Resources in the Bering Sea/Aleutian Islands Region as Projected for 2021, Chapter 4. North Pacific Fishery Management Council, P. O. Box 103136, Anchorage, AK 99510.

Wakabayashi, K. 1989. Studies on the fishery biology of yellowfin sole in the eastern Bering Sea. [In Jpn., Engl. Summ.] Bull. Far Seas Fish. Res. Lab. 26:21-152.

Yeung, C., and Yang, M.-S. 2018. Spatial variation in habitat quality for juvenile flatfish in the southeastern Bering Sea and its implications for productivity in a warming ecosystem. Journal of Sea Research, 139: 62–72.

Yeung, C. and Cooper, D.W., 2019. Contrasting the variability in spatial distribution of two juvenile flatfishes in relation to thermal stanzas in the eastern Bering Sea. ICES Journal of Marine Science, 77(3), pp.953-963.

Yeung, C., Copeman, L.A., Matta, M.E. and Yang, M.S., 2021. Latitudinal variation in the growth and condition of Juvenile flatfishes in the Bering Sea. Estuarine, Coastal and Shelf Science, 258, p.107416.

\pagebreak

# Figures

```{r agetmp, message=FALSE, echo=FALSE, warning=FALSE}

OUT_len3=data.frame(cbind(rep(OUT_len2$Age1,2),c(OUT_len2$Growth_Male,OUT_len2$Growth_Female),c(rep("Male",nrow(OUT_len2)),rep("Female",nrow(OUT_len2)))))
colnames(OUT_len3)=c("Age","Growth","Sex")
OUT_len3$Age=as.numeric(OUT_len3$Age)
OUT_len3$Sex=as.factor(OUT_len3$Sex)
OUT_len3$Growth=as.numeric(OUT_len3$Growth)
#offset the Age by .5 for males
OUT_len3$Age[which(OUT_len3$Sex=="Male")]=OUT_len3$Age[which(OUT_len3$Sex=="Male")]+0.2
colnames(OUT_len3)=c("Age","Growth","Sex")
ggplot(OUT_len3)+geom_point(aes(x=Age,y=Growth,color=Sex),alpha=.9,size=2)+theme_bw()+ggtitle("YFS Growth by Age")+ylab("Growth (cm)")+xlab("First age of growth increment")+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+theme(axis.title=element_text(size=14))

#ggplot(data=OUT_len2)+geom_point(aes(x=Age,y=Growth_Male))+theme_bw()+ggtitle("Male Growth by Age")#Year implies year of birth only

#ggplot(data=OUT_len2)+geom_point(aes(x=Age1,y=Growth_Female))+theme_bw()+ggtitle("Female Growth by Age")#Year implies year of birth only


#ggplot(data=OUT_len2)+geom_point(aes(x=Age1,y=Growth_Male,color=Temperature))+theme_bw()+ggtitle("Male Growth by Age")+xlab("First age of the growth increment")#Year implies year of birth only

#ggplot(data=OUT_len2)+geom_point(aes(x=Age1,y=Growth_Female,color=Temperature))+theme_bw()+ggtitle("Female Growth by Age")+xlab("First age of the growth increment")#Year implies year of birth only

#I do not think that we should look at variation in growth by year
# Because Year implies year class rather than year of growth
# and later year classes have fewer observations.
# We could say Year<2005
#ggplot(data=OUT_len2[which(OUT_len2$Year<2005),])+geom_point(aes(x=Year,y=Growth_Male,color=Year))+theme_bw()+ggtitle("Male Growth by Age")+xlab("Year class")#Year implies year of birth only

#ggplot(data=OUT_len2)+geom_point(aes(x=Year,y=Growth_Female,color=Year))+theme_bw()+ggtitle("Female Growth by Age")+xlab("Year Class")#Year implies year of birth only

#leveneTest(Growth_Female~Age_factor,OUT_len2)

#leveneTest(Growth_Male~Age_factor,OUT_len2)

#leveneTest(Growth_Female~Age_factor,OUT_len2[which(OUT_len2$Age1<13),])

#leveneTest(Growth_Male~Age_factor,OUT_len2[which(OUT_len2$Age1<12),])

```

`r figure_nums(name="agetmp",caption = "Male and female Yellowfin Sole growth by age, 1982-2019. Note that males and females are slightly offset.")`

\pagebreak

```{r gtemp, echo=FALSE, message=FALSE, warning=FALSE}
#Growth by length for all ages and temperature (year too but year corelates with temp)

gt_M=ggplot(data=OUT_len2[which(OUT_len2$Age1<12),])+geom_point(aes(x=Temperature,y=Growth_Male,color=Age1))+theme_bw()+
 ggtitle("YFS male growth increment up to age 11 to 12, \nby increment and temperature")+ylim(-3,9)+
 ylab("Growth (cm)")+
 guides(col= guide_legend(title= "Growth\nincrement\nfirst age"))


gt_F=ggplot(data=OUT_len2[which(OUT_len2$Age1<13),])+geom_point(aes(x=Temperature,y=Growth_Female,color=Age1))+
 guides(col= guide_legend(title= "First age\nof growth\nincrement"))+ylim(-3,9)+xlab("Temperature (°C)")+
 ggtitle("YFS female growth increment up to age 12-13, \nby increment and temperature")+
 ylab("Growth (cm)")+theme_bw()+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=14))+theme(axis.title=element_text(size=15))

gt_M=ggplot(data=OUT_len2[which(OUT_len2$Age1<13),])+geom_point(aes(x=Temperature,y=Growth_Male,color=Age1))+
 guides(col= guide_legend(title= "First age\nof growth\nincrement"))+ylim(-3,9)+xlab("Temperature (°C)")+
 ggtitle("YFS male growth increment up to age 11-12, \nby increment and temperature")+
 ylab("Growth (cm)")+theme_bw()+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=14))+theme(axis.title=element_text(size=15))

grid.arrange(gt_M,gt_F)
```

`r figure_nums(name="gtemp",caption = "Growth increments (mm) by male and female Yellowfin sole with respect to <100 m bottom temperature in the Bering Sea. Increments are shown from age 2 to 3 years through 12 to 13 for females and 11 to 12 for males. ")`

\pagebreak

```{r coldpooltmp,echo=FALSE,message=FALSE}
temCP=data.frame(rep(seq(1982,2019,1),2),c(tempz$AVGBSBT[1:38],(-.1*((cold_pool_index$AREA_LTE2_KM2[1:38]/15000)+10)+5)),c(rep("Bottom_Temp",38),rep("Cold_Pool_Index",38)))
colnames(temCP)=c("Year","Measurement","Type")
ggplot(data=temCP)+geom_line(aes(x=Year,y=Measurement,color=Type))+theme_bw()+ggtitle("Cold Pool and Bottom Temperature are very similar indices")

#plot(seq(1982,2019,1),tempz$AVGBSBT[1:38],type="l")
#lines(seq(1982,2019,1),.1*((cold_pool_index$AREA_LTE2_KM2[1:38]/15000)+10),col="red")

```

`r figure_nums(name="coldpooltmp",caption = "Comparison of the cold pool index (LTE2_KM2) and bottom temperature, 1982 - 2019. Note: cold pool index is multiplied by negative one for comparison.")`

\pagebreak

```{r nesseplot,echo=FALSE,message=FALSE}
Mal.lm<-lm(Growth_Male~Age1+Temperature,data=OUT_len2[which(OUT_len2$Age1<12),])

Fem.lm<-lm(Growth_Female~Age1+Temperature,data=OUT_len2[which(OUT_len2$Age1<13),])

plot(OUT_len2[which(OUT_len2$Age1==4),]$Temperature,OUT_len2[which(OUT_len2$Age1==4),]$Growth_Male,pch=as.character(OUT_len2[which(OUT_len2$Age1==4),]$Age_factor),main="Male Growth",xlab="Temperature",ylab="Growth (mm)")
temp=seq(0,5,1)
lines(temp,Mal.lm$coefficients[1]+1*Mal.lm$coefficients[2]+temp*Mal.lm$coefficients[3])

#now do same with Temperature anomalies because that is what is used in the data

OUT_len2$Tempanom=OUT_len2$Temperature-mean(OUT_len2$Temperature)

Mal_anom.lm<-lm(Growth_Male~Age1+Tempanom,data=OUT_len2[which(OUT_len2$Age1<12),])

Fem_anom.lm<-lm(Growth_Female~Age1+Tempanom,data=OUT_len2[which(OUT_len2$Age1<13),])

plot(OUT_len2[which(OUT_len2$Age1==4),]$Tempanom,OUT_len2[which(OUT_len2$Age1==4),]$Growth_Male,pch=as.character(OUT_len2[which(OUT_len2$Age1==4),]$Age_factor),main="Male Growth",xlab="Temperature Anomaly (1982-201?",ylab="Growth (mm)")
temp=seq(0,5,1)
lines(seq(-2.5,2.5),Mal_anom.lm$coefficients[1]+1*Mal_anom.lm$coefficients[2]+seq(-2.5,2.5)*Mal_anom.lm$coefficients[3])


```

#try plotting another way

# Tables


```{r levene,echo=FALSE,message=FALSE }

leveneTest(Growth_Female~Age_factor,OUT_len2)

leveneTest(Growth_Male~Age_factor,OUT_len2)

```
`r table_nums(name="levene",caption = "Levene's test for homogeneity of growth variance by age.")`

\pagebreak

```{r ANOVA,echo=FALSE,message=FALSE,warning=FALSE}


#summary(Fem.lm)
fem1=data.frame(summary(Fem.lm)$coefficients)
fem1[,1:3]=round(fem1[,1:3],3)
fem1[,4]=signif(fem1[,4],digits=4)
colnames(fem1)[2]="Std.Error"
colnames(fem1)[4]="Pr(t)"

#fit2=aov(Growth_Female~Temperature+Age1,OUT_len2[which(OUT_len2$Age1<13),])
#Anova(fit2, type="III")


mal1=data.frame(summary(Mal.lm)$coefficients)
mal1[,1:3]=round(mal1[,1:3],3)
mal1[,4]=signif(mal1[,4],digits=4)
colnames(mal1)[2]="Std.Error"
colnames(mal1)[4]="Pr(t)"



kable(mal1,"latex",booktabs=T,linesep="",longtable=FALSE,align='r',digits=35)%>%
  add_header_above(c("Male"=5))

kable(fem1,"latex",booktabs=T,linesep="",longtable=FALSE,align='r',digits=35)%>%
  add_header_above(c("Female"=5))


#In code we have Growth_Option =1 is just no temperature effect.
#Growth_Option>1 is 

```

`r table_nums(name="anova",caption = "Anova results for male and female Yellowfin Sole, comparing the effect of bottom temperature and age on growth increment. Age and temperature interactions were not significant and were not included in the final anova.")`


```{r tempgrowth,echo=FALSE,message=FALSE, warning=FALSE}

#switch to cm.
#There were 16,431 males and 21,050 females
> dim(srv_wtage[which(srv_wtage$SEX<3&srv_wtage$LENGTH>0&srv_wtage$AGE>0&srv_wtage$SEX==1),])
[1] 16431    21
> dim(srv_wtage[which(srv_wtage$SEX<3&srv_wtage$LENGTH>0&srv_wtage$AGE>0&srv_wtage$SEX==2),])
[1] 21050    21

#Male
VBf_M=growth(intype=1,unit=1,size=srv_wtage$LENGTH[which(srv_wtage$SEX==1)]/10,age=srv_wtage$AGE[which(srv_wtage$SEX==1)],calctype=1,wgtby=1,error=1,Sinf=50,K=0.13,t0=0.3)
P_M=summary(VBf_M$vout)$parameters[,1]
age=seq(1,20,1);Sinf=P_M[1];K=P_M[2];t0=P_M[3]
Males=Sinf*(1-exp(-(K*(age-t0))))

#Female
VBf_F=growth(intype=1,unit=1,size=srv_wtage$LENGTH[which(srv_wtage$SEX==2)]/10,age=srv_wtage$AGE[which(srv_wtage$SEX==2)],
           calctype=1,wgtby=1,error=1,Sinf=50,K=0.13,t0=0.3)
P_F=summary(VBf_F$vout)$parameters[,1]
age=seq(1,20,1);Sinf=P_F[1];K=P_F[2];t0=P_F[3]
Females=Sinf*(1-exp(-(K*(age-t0))))

#calculate temperature deviations
tempanom=tempz$AVGBSBT-mean(tempz$AVGBSBT)
base_incr_f=vector()
base_incr_m=vector()
for(i in 1:19){
base_incr_f[i]=Females[i+1]-Females[i]
base_incr_m[i]=Males[i+1]-Males[i]
}

Year=seq(1981,2022,1)
X=length(Year)-1

Fem_gr=matrix(0,length(Year),20)
for(i in 1:X){
Fem_gr[i+1,c(2:14)]=Fem_gr[i,c(1:13)]+base_incr_f[1:13]+.24*tempanom[i]}#first yaer of tempanom is 1982 and matrix starts in 1981
rownames(Fem_gr)=Year
Fem_gr[1,]=Females


#Here for males do one age lower just through 12.
Mal_gr=matrix(0,length(Year),20)
for(i in 1:X){
Mal_gr[i+1,c(2:14)]=Mal_gr[i,c(1:13)]+base_incr_m[1:13]+.26*tempanom[i]#first yaer of tempanom is 1982 and matrix starts in 1981
#Fill in blank years
Mal_gr[i+1,1]=Males[1]
Mal_gr[i+1,15:20]=Mal_gr[i,14:19]+base_incr_m[14:19]
}

#compare with lenM and LenF
#To use same code add 10 more rows to top.
Fem_gr1=data.frame(rbind(matrix(0,10,20),Fem_gr))
rownames(Fem_gr1)=seq(1971,2022,1)

Mal_gr1=data.frame(rbind(matrix(0,10,20),Mal_gr))
rownames(Mal_gr1)=seq(1971,2022,1)

N=c(rep(18,18),18,18,rev(seq(2,17,1)))#This represents the number of years a cohort is represented by the data  and removed 2020

Year=seq(1982,2018,1)
matM_lenA=matrix(0,length(Year),length(seq(2,19,1)))
matF_lenA=matrix(0,length(Year),length(seq(2,19,1)))
tempMAT_lenA=matrix(0,length(Year),length(seq(2,19,1)))
colnames(matM_lenA)=seq(2,19,1);colnames(matF_lenA)=seq(2,19,1)
rownames(matF_lenA)=Year;rownames(matM_lenA)=Year
for(j in 1:length(N)){
 for(i in 1:N[j]){
  matF_lenA[j,i]=Fem_gr1[(9+i+j+2),(i+2)]-Fem_gr1[(9+i+j+1),(i+1)]
  matM_lenA[j,i]=Mal_gr1[(9+i+j+2),(i+2)]-Mal_gr1[(9+i+j+1),(i+1)]
  tempMAT_len[j,i]=tempz[which(tempz$YEAR==1982+i+j-1),3]
 }}

colnames(tempMAT_lenA)=seq(2,19,1);colnames(matM_lenA)=seq(2,19,1);colnames(matF_lenA)=seq(2,19,1)

matF_lenB=cbind(Year,matF_lenA);matM_lenB=cbind(Year,matM_lenA)

matF2_lenB=reshape2::melt(matF_lenB,id=c("Year"));matM2_lenB=reshape2::melt(matM_lenB,id=c("Year"))
matF2_lenC=matF2_lenB[38:nrow(matF2_lenB),];matM2_lenC=matM2_lenB[38:nrow(matM2_lenB),]
tempMAT2_lenA=reshape2::melt(tempMAT_lenA,id=c("Year"));
colnames(tempMAT2_lenA)=c("Year","Age1","Temperature")
colnames(matF2_lenC)=c("Year","Age1","Growth_Female");colnames(matM2_lenC)=c("Year","Age1","Growth_Male")
OUT_lenA=cbind(tempMAT2_len,matF2_lenC$Growth_Female,matM2_lenC$Growth_Male)
colnames(OUT_lenA)=c("Year","Age1","Temperature","Growth_Female","Growth_Male")

OUT_lenA2=OUT_lenA[which(OUT_lenA$Temperature>0),]
#Look at the data
#Males and females grow in length more early in life
OUT_lenA2$Age_factor=as.factor(OUT_lenA2$Age1)#set up age as a factor

#What is the maximum length and age for males and females?

max(srv_wtage$LENGTH[which(srv_wtage$SEX==1)]) #Males are 1?
[1] 490
max(srv_wtage$LENGTH[which(srv_wtage$SEX==2)])
[1] 500
max(srv_wtage$AGE[which(srv_wtage$SEX==2)],na.rm=TRUE)
[1] 38
max(srv_wtage$AGE[which(srv_wtage$SEX==1)],na.rm=TRUE)
[1] 43


library(ggpmisc)
```

```{r quad,echo=FALSE,message=FALSE, warning=FALSE}

#now try quadratic - the fit is good
library(ggpubr)
meanz=data.frame(cbind(rep(mean(tempz$AVGBSBT),length(base_incr_m))),base_incr_m,base_incr_f);colnames(meanz)=c("Temp","AvgGrowth_m","AvgGrowth_f")

#Figure Female Quadratic
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(OUT_len2[which(OUT_len2$Age1<12),], aes(x=Temperature, y=Growth_Female)) +ylim(c(-3,9))+xlab("Temperature (°C)")+
  geom_point() +
  stat_smooth(method = "lm", formula = formula) +stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),col="grey40",formula = formula, size = 7
  ) +ylab("Female Growth Increments (cm)") + theme_bw()+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+ggtitle("Female Quadratic Fit")+theme(axis.title=element_text(size=15))

#+geom_point(data=meanz,aes(x=Temp,y=AvgGrowth_f),col="red")

#Figure Male Quadratic
formula <- y ~ poly(x, 2, raw = TRUE)
ggplot(OUT_len2[which(OUT_len2$Age1<12),], aes(x=Temperature, y=Growth_Male)) +ylim(c(-3,9))+xlab("Temperature (°C)")+
  geom_point() +
  stat_smooth(method = "lm", formula = formula) +stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),col="grey40",formula = formula, size = 7
  ) +ylab("Male Growth Increments (cm)") + theme_bw()+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+ggtitle("Male Quadratic Fit")+theme(axis.title=element_text(size=15))

#geom_point(data=meanz,aes(x=Temp,y=AvgGrowth_m),col="red")+

#R squared is teh proportion of variation in teh dependent variable that is predictable from the independent variable.

```



```{r linear,echo=FALSE,message=FALSE, warning=FALSE}
formula <- y ~ x
ggplot(OUT_len2[which(OUT_len2$Age1<13),], aes(x=Temperature, y=Growth_Female)) +ylim(-3,9)+xlab("Temperature (°C)")+
  geom_point() +
  stat_smooth(method = "lm", formula = formula) +stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),col="grey40",formula = formula,size=7
  ) +ylab("Female Growth Increments (cm)")+theme_bw()+
 theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+ggtitle("Female Linear Fit")+theme(axis.title=element_text(size=15))
#note that it is better when you restrict older ages.

formula <- y ~ x
ggplot(OUT_len2[which(OUT_len2$Age1<12),], aes(x=Temperature, y=Growth_Male)) +ylim(-3,9)+xlab("Temperature (°C)")+
  geom_point() +
  stat_smooth(method = "lm", formula = formula) +stat_regline_equation(aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),col="grey40",formula = formula,size=7
  ) +ylab("Male Growth Increments (cm)")+theme_bw()+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14))+ggtitle("Male Linear Fit")+theme(axis.title=element_text(size=15))
```

This fits okay but somehow not great at the ends. Somehow my idea is not fitting as well as it seemse like it should.

```{r combine_empirical,echo=FALSE,message=FALSE, warning=FALSE}
#maybe combine the two
#Figure 5 Female linear
ggplot()+geom_point(data=OUT_len2[which(OUT_len2$Age<13),],aes(x=Temperature,y=Growth_Female,color=Age))+
new_scale_colour()+
geom_point(data=OUT_lenA2[which(OUT_lenA2$Age<13),],aes(x=Temperature+.05,y=Growth_Female,color=Age))+
scale_colour_gradient(low = "red", high = "orange")+
 guides(col= guide_legend(title= "First age\nof growth\nincrement"))+
 ggtitle("Empirical female growth (blues) and modeled linear fit (reds)")+ylab("Growth (cm)")+theme_bw()+xlab("Temperature (°C)")+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size =14),title=element_text(size=12))+theme(axis.title=element_text(size=15))

#model after this one
ggplot()+geom_point(data=OUT_len2[which(OUT_len2$Age<12),],aes(x=Temperature,y=Growth_Male,color=Age))+
new_scale_colour()+
geom_point(data=OUT_lenA2[which(OUT_lenA2$Age<12),],aes(x=Temperature+.05,y=Growth_Male,color=Age))+
scale_colour_gradient(low = "red", high = "orange")+
 guides(col= guide_legend(title= "First age\nof growth\nincrement"))+
 ggtitle("Empirical male growth (blues) and modeled quadratic fit (reds)")+
 ylab("Growth (mm)")+theme_bw()+xlab("Temperature (°C)")+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))


#males
ggplot()+geom_point(data=OUT_len2[which(OUT_len2$Age1<12),],aes(x=Temperature,y=Growth_Male,color=Age1))+
new_scale_colour()+
geom_point(data=OUT_lenA2[which(OUT_lenA2$Age1<12),],aes(x=Temperature+.05,y=Growth_Male,color=Age1))+
scale_colour_gradient(low = "red", high = "orange")+
 guides(col= guide_legend(title= "Modeled growth\nincrement\nfirst age"))+
 ggtitle("YFS male linear growth increment up to age 11 to 12, \nby increment and temperature\nModeled (reds) and empirical (blues)")+
 ylab("Growth (mm)")+theme_bw()+xlab("Temperature")
```

```{r combine_quadratic,echo=FALSE,message=FALSE, warning=FALSE}
#calculate temperature deviations
tempanom=tempz$AVGBSBT-mean(tempz$AVGBSBT)
Year=seq(1981,2022,1)
Fem_gr=matrix(0,length(Year),20)
X=length(Year)-1

for(i in 1:X){
Fem_gr[i+1,c(2:14)]=Fem_gr[i,c(1:13)]+base_incr_f[1:13]+.35*tempanom[i]-.023*tempanom[i]^2}#first yaer of tempanom is 1982 and matrix starts in 1981
rownames(Fem_gr)=Year
Fem_gr[1,]=Females
Mal_gr[1,]=Males
base_incr_f=vector()
base_incr_m=vector()
for(i in 1:19){
base_incr_f[i]=Females[i+1]-Females[i]
base_incr_m[i]=Males[i+1]-Males[i]
}

#Here for males do one age lower just through 12.

Year=seq(1981,2022,1)
X=length(Year)-1
Mal_gr=matrix(0,length(Year),20)
for(i in 1:X){
Mal_gr[i+1,c(2:14)]=Mal_gr[i,c(1:13)]+base_incr_m[1:13]+.69*tempanom[i]-.069*tempanom[i]^2#first yaer of tempanom is 1982 and matrix starts in 1981
#Fill in blank years
Mal_gr[i+1,1]=Males[1]
Mal_gr[i+1,15:20]=Mal_gr[i,14:19]+base_incr_m[14:19]
}

#compare with lenM and LenF
#To use same code add 10 more rows to top.
Fem_gr1=data.frame(rbind(matrix(0,10,20),Fem_gr))
rownames(Fem_gr1)=seq(1971,2022,1)

Mal_gr1=data.frame(rbind(matrix(0,10,20),Mal_gr))
rownames(Mal_gr1)=seq(1971,2022,1)

N=c(rep(18,18),18,18,rev(seq(2,17,1)))#This represents the number of years a cohort is represented by the data  and removed 2020

Year=seq(1982,2018,1)
matM_lenA=matrix(0,length(Year),length(seq(2,19,1)))
matF_lenA=matrix(0,length(Year),length(seq(2,19,1)))
tempMAT_lenA=matrix(0,length(Year),length(seq(2,19,1)))
colnames(matM_lenA)=seq(2,19,1);colnames(matF_lenA)=seq(2,19,1)
rownames(matF_lenA)=Year;rownames(matM_lenA)=Year
for(j in 1:length(N)){
 for(i in 1:N[j]){
  matF_lenA[j,i]=Fem_gr1[(9+i+j+2),(i+2)]-Fem_gr1[(9+i+j+1),(i+1)]
  matM_lenA[j,i]=Mal_gr1[(9+i+j+2),(i+2)]-Mal_gr1[(9+i+j+1),(i+1)]
  tempMAT_len[j,i]=tempz[which(tempz$YEAR==1982+i+j-1),3]
 }}

colnames(tempMAT_lenA)=seq(2,19,1);colnames(matM_lenA)=seq(2,19,1);colnames(matF_lenA)=seq(2,19,1)

matF_lenB=cbind(Year,matF_lenA);matM_lenB=cbind(Year,matM_lenA)

matF2_lenB=reshape2::melt(matF_lenB,id=c("Year"));matM2_lenB=reshape2::melt(matM_lenB,id=c("Year"))
matF2_lenC=matF2_lenB[38:nrow(matF2_lenB),];matM2_lenC=matM2_lenB[38:nrow(matM2_lenB),]
tempMAT2_lenA=reshape2::melt(tempMAT_lenA,id=c("Year"));
colnames(tempMAT2_lenA)=c("Year","Age1","Temperature")
colnames(matF2_lenC)=c("Year","Age1","Growth_Female");colnames(matM2_lenC)=c("Year","Age1","Growth_Male")
OUT_lenA=cbind(tempMAT2_len,matF2_lenC$Growth_Female,matM2_lenC$Growth_Male)
colnames(OUT_lenA)=c("Year","Age1","Temperature","Growth_Female","Growth_Male")

OUT_lenA2=OUT_lenA[which(OUT_lenA$Temperature>0),]
#Look at the data
#Males and females grow in length more early in life
OUT_lenA2$Age_factor=as.factor(OUT_lenA2$Age1)#set up age as a factor

ggplot()+geom_point(data=OUT_len2[which(OUT_len2$Age1<13),],aes(x=Temperature,y=Growth_Female,color=Age1))+
new_scale_colour()+
geom_point(data=OUT_lenA2[which(OUT_lenA2$Age1<13),],aes(x=Temperature+.05,y=Growth_Female,color=Age1))+
scale_colour_gradient(low = "red", high = "orange")+
 guides(col= guide_legend(title= "Modeled growth\nincrement\nfirst age"))+
 ggtitle("YFS female growth quadratic up to age 12 to 13, \nby increment and temperature modeled (reds) \nand empirical (blues)")+ylab("Growth (cm)")+theme_bw()+xlab("Temperature")+geom_point(data=meanz[1:11,],aes(x=Temp,y=AvgGrowth_m),col="green")+geom_point(data=meanz[1:11,],aes(x=Temp-2,y=AvgGrowth_m-2*(.26)),col="green")+geom_point(data=meanz[1:11,],aes(x=Temp+2,y=AvgGrowth_m+2*(.26)),col="green")+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=14))+theme(axis.title=element_text(size=15))

colnames(OUT_lenA2)[2]="Age"
colnames(OUT_len2)[2]="Age"

#Figure 5 male quadratic
ggplot()+geom_point(data=OUT_len2[which(OUT_len2$Age<12),],aes(x=Temperature,y=Growth_Male,color=Age))+ylim(-3,9)+
new_scale_colour()+
geom_point(data=OUT_lenA2[which(OUT_lenA2$Age<12),],aes(x=Temperature+.05,y=Growth_Male,color=Age))+
scale_colour_gradient(low = "red", high = "orange")+
 guides(col= guide_legend(title= "First age\nof growth\nincrement"))+
 ggtitle("Empirical male growth (blues) and modeled quadratic fit (reds)")+
 ylab("Growth (mm)")+theme_bw()+xlab("Temperature (°C)")+theme(axis.text.x = element_text(size = 14),axis.text.y = element_text(size = 14),title=element_text(size=12))+theme(axis.title=element_text(size=15))
```
#do for females through age 13 to 14

```{r ssq, echo=FALSE,message=FALSE}
OUT_len2$Temperature2=OUT_len2$Temperature^2
quad_F=lm(formula = Growth_Female ~ Temperature + Temperature2, data = OUT_len2[which(OUT_len2$Age1<13),])
summary(quad_F)
#Residual standard error: 1.372 on 318 degrees of freedom
#Multiple R-squared:  0.04005,	Adjusted R-squared:  0.03401
linear_F=lm(formula = Growth_Female ~ Temperature, data = OUT_len2[which(OUT_len2$Age1<13),])
summary(linear_F)
#Residual standard error: 1.37 on 319 degrees of freedom
#Multiple R-squared:  0.03941,	Adjusted R-squared:  0.0364 

OUT_len2$Temperature2=OUT_len2$Temperature^2
quad_M=lm(formula = Growth_Male ~ Temperature + Temperature2, data = OUT_len2[which(OUT_len2$Age1<12),])
summary(quad_M)
#Residual standard error: 1.271 on 287 degrees of freedom
#Multiple R-squared:  0.06274,	Adjusted R-squared:  0.05621
linear_M=lm(formula = Growth_Male ~ Temperature, data = OUT_len2[which(OUT_len2$Age1<12),])
summary(linear_M)
#Residual standard error: 1.274 on 288 degrees of freedom
#Multiple R-squared:  0.05603,	Adjusted R-squared:  0.05275


#quadratic lowers residual error and increases R2 in males.  quadratic better males
#quadratic increases residual error and lowers R2 in females. linear better females

```

```


```{r wtlen, echo=FALSE,message=FALSE}
#For ADMB code it is in weight so I'll convert to weights first.

#	"a,"	"b,"	Females																																																										
#	female	l-w	relationship	wt(g)=0.0059*L(cm)^3.205		  #updated 2021 

#here Females represents Female length at age.
Female_wt=0.0059*Females^3.205																																																																																																																		
#	male	l-w	relationship	wt(g)=0.0091*L(cm)^3.068		   #updated 2021 																																																						
#	"a,"	"b,"	Males																																																										
Male_wt=0.0091*Males^3.068	

basewt_incr_f=vector()
basewt_incr_m=vector()
for(i in 1:19){
basewt_incr_f[i]=Female_wt[i+1]-Female_wt[i]
basewt_incr_m[i]=Male_wt[i+1]-Male_wt[i]
}

Year=seq(1981,2022,1)
X=length(Year)-1
Mal_gr=matrix(0,length(Year),20)
Mal_gr[1,]=Male_wt
for(i in 1:X){
Mal_gr[i+1,1]=Male_wt[1]
Mal_gr[i+1,c(2:14)]=Mal_gr[i,c(1:13)]+basewt_incr_m[1:13]+.69*tempanom[i]-.069*tempanom[i]^2

Mal_gr[i+1,15:20]=Mal_gr[i,14:19]+basewt_incr_m[14:19]
}

#next put these into total weight by age and year a cumulative summation

#first yaer of tempanom is 1982 and matrix starts in 1981
#Fill in blank years

#Can you apply the same change in weight as length? Yes, does not matter so just apply pattern to weights not lengths.
#Hmm, maybe it does matter - I added parenthesis to the x here.
x=5
wt1=.7*2*(x^1.3)
wt2=2*((.7*x)^1.3)

#try the shaded boxes plot

Femz=YFS18_2$oac_srv_s[4:41,1:20]
Malz=YFS18_2$oac_srv_s[4:41,21:40]

yrs=length(YFS18_2$yrs_srv_age_s[4:41])
Age=rep(seq(1,20,1),each=yrs)
Frequency=c(Femz)
Year=rep(YFS18_2$yrs_srv_age_s[4:41],length(yrs))
fishlen=data.frame(Year,Age,Frequency)
Fem <- ggplot(fishlen, aes(Age,Year)) +
 geom_tile(aes(fill = Frequency),colour = "lightgrey")+
 scale_fill_gradient(low = "white",high = "darkblue")+ 
 scale_y_continuous(breaks=seq(1979,2021,2),labels=seq(1979,2021,2))+ 
 ggtitle("YFS Ages - Survey Females") +
 theme(axis.text.y=element_text(size=rel(.7)))+theme_bw()

Frequency=c(Malz)
fishlen=data.frame(Year,Age,Frequency)
Mal <- ggplot(fishlen, aes(Age,Year)) +
 geom_tile(aes(fill = Frequency),colour = "lightgrey")+
 scale_fill_gradient(low = "white",high = "darkblue")+ 
 scale_y_continuous(breaks=seq(1975,2021,2),labels=seq(1975,2021,2))+ 
 ggtitle("YFS Ages - Survey Males") +
 theme(axis.text.y=element_text(size=rel(.7)))+theme_bw()

grid.arrange(Fem,Mal,ncol=1)



#Why is the population declining?Is it natural mortality?
#fishiing mortality is pretty low, and not really increasing
plot(seq(1982,2021),YFS18_2$F_f[29:68,20],ylim=c(0,.11),type="l")

#survey biomass has declined since 1092 and 1983 but isn't really declining much..
plot(YFS18_2$yrs_srv,YFS18_2$ob_bts,type="l")

plot(YFS18_2$yrs_srv,YFS18_2$env_cov[1,],type="l")#temperature goes up and down

plot(YFS18_2$yrs_srv,YFS18_2$env_cov[2,],type="l")#survy start date anomalies not really different except around the year2000.

#
Year=seq(1954,2021)
natage_pf=data.frame(cbind(Year,YFS18_2$natage_f))
colnames(natage_pf)=c("Year",seq(1,20,1))
natage_pf2=reshape2::melt(natage_pf,id="Year")
colnames(natage_pf2)=c("Year","Age","Frequency")


freqsF=natage_pf2[which(natage_pf2$Year>2018),]
freqsF$Age=as.numeric(freqsF$Age)
freqsF$Year=as.factor(freqsF$Year)
sapply(freqsF,class)
ggplot(freqsF)+geom_line(aes(x=Age,y=Frequency,col=Year))

```

